---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "70%"
)
```

<!-- badges: start -->
<!-- badges: end -->

```{r setup, eval=FALSE, include=FALSE, warning=FALSE, message=FALSE}
#devtools::install_github("gestimation/cifmodeling") 
#install.packages(c("Rcpp","nleqslv","boot","ggplot2","ggsurvfit","patchwork","modelsummary"))
if (requireNamespace("cifmodeling", quietly = TRUE)) {
  library(cifmodeling)
} else {
  stop("Package 'cifmodeling' is not available during vignette build.")
}
library(nleqslv)
library(boot)
library(ggplot2)
library(ggsurvfit)
library(patchwork)
library(modelsummary)
data("diabetes.complications", package = "cifmodeling")
set.seed(1)
```

# Visualization and Modeling of Competing Risks in R — cifmodeling

## Quick start

This package is a compact and unified toolkit for the Kaplan–Meier/Aalen–Johansen curves, visualization, and direct polytomous regression for survival and competing risks data. 

```{r example01-1, fig.cap="Aalen–Johansen cumulative incidence curves from cifplot()", warning=FALSE, message=FALSE}
library(cifmodeling)
data(diabetes.complications)
cifplot(Event(t,epsilon)~fruitq, data=diabetes.complications, 
        outcome.type="competing-risk", panel.per.event=TRUE)
```

This is an example code snippet applying `cifplot()` to visualize the Aalen–Johansen cumulative incidence functions (CIFs). In competing risks data, censoring is often coded as 0, events of interest as 1, and competing risks as 2. The variable `epsilon` in `diabetes.complications` data frame represents the occurrence of competing risks according to this coding scheme. By setting `panel.per.event=TRUE`, the CIF curve for diabetic retinopathy (`epsilon=1`) is output on the left and macrovascular complications (`epsilon=2`) on the right. 

## Tools for survival and competing risks analysis

In clinical and epidemiological research, analysts often need to handle censoring, competing risks, and intercurrent events (e.g., treatment switching), but existing R packages typically separate these tasks into different interfaces. `cifmodeling` provides a **unified, publication-ready toolkit** that integrates description of survival and CIF curves, regression modeling, and visualization for survival and competing risks data. It covers both nonparametric estimation and regression modeling of CIFs, centered around three tightly connected functions.

-   `cifplot()` typically generates a survival or CIF curve with marks that represent censoring, competing risks and intercurrent events. Multiple standard error (SE) estimators and confidence interval (CI) methods are supported. Visualization relies on `ggsurvfit/ggplot2`. 

-   `cifpanel()` generates a multi-panel figure for survival/CIF curves, arranged either in a grid layout or as an inset overlay.

-   `polyreg()` fits coherent regression models of CIFs based on polytomous log odds products and the stratified inverse probability of censoring weighting (IPCW) estimator. This function is particularly well-suited for causal inference in terms of typical effect measures, namely risk ratios, odds ratios, and subdistribution hazard ratios, with a competing risks, survival, or binary outcome. 

These functions adopt a formula + data syntax, return tidy, publication-ready outputs, and integrate seamlessly with `ggsurvfit` and `modelsummary` for visualization.

## Installation

The package is implemented in R and relies on `Rcpp`, `nleqslv` and `boot` for the numerical back-end. The examples in this document also use `ggplot2`, `ggsurvfit`, `patchwork` and `modelsummary` for tabulation and plotting. Install the core package and these companion packages with:

``` r
install.packages("Rcpp")
install.packages("nleqslv")
install.packages("boot")
install.packages("ggplot2")
install.packages("ggsurvfit")
install.packages("patchwork")
install.packages("modelsummary")
devtools::install_github("gestimation/cifmodeling")
```

## Interoperability

There are several excellent R packages for survival and competing-risk analysis. The `survival` package provides the canonical API for survival data. In combination with `ggsurvfit`, `survival::survfit()` can generate publication-ready survival plots. For CIF plots, integration in the general ecosystem is not as streamlined; `cifmodeling` fills this gap by offering `cifplot()` for survival/CIF plots and multi-panel figures via a single, unified interface.

If you need fine-grained plot customization, compute the estimator and keep a survfit-compatible object with `cifcurve()` (or supply your own survfit object) and then style it using `ggsurvfit/ggplot2` layers. In other words, use `cifcurve()` for estimation, `cifplot()/cifpanel()` for quick high-quality figures, and fall back to the `ggplot` ecosystem when you want full artistic control.

For interval estimation with unweighted survival data, `cifcurve()` mirrors the CI options of `survival::survfit()` and the CI based on influence functions additionally  (`error="greenwood"`, `"tsiatis"` and `"if"`). For weighted survival data, `cifcurve()` adds valid CIs and SEs of Xie and Liu (2005), and for competing risk data, `error="delta"`, `"aalen"` and `"if"` options. The defaults in `cifmodeling` are chosen from external and internal simulation evidence; they should be sensible for most applications, while still allowing experts to override them when needed. 

The `mets` package is a more specialized toolkit that provides advanced statistical methods for competing risk analysis. `cifmodeling::polyreg()` focuses on coherent modeling of all CIFs simultaneously to estimate RR/OR/SHR at user-specified times; this coherence can come with longer runtimes on large problems. If you prefer fitting separate regression models for each competing event or need Fine–Gray specifically, `mets::cifreg()` (and `mets::binreg()` for RR and OR estimation) are excellent choices. 

## Quality control

`cifmodeling` includes an extensive test suite built with **testthat**, ensuring the numerical accuracy and graphical consistency of all core functions (`cifcurve`, `cifplot`, `cifpanel`, and `polyreg`). The package is continuously tested on GitHub Actions (Windows, macOS, Linux) to maintain reproducibility and CRAN-level compliance.

## Main functions

### cifplot()

This function produces the **Kaplan–Meier survival** or **Aalen–Johansen cumulative incidence** curve from a unified formula + data interface. It auto-labels axes based on `outcome.type` and `type.y` (`"risk"` for 1-survival or CIF), can add censoring/competing-risk/intercurrent-event marks, and returns a regular ggplot object (compatible with + and %+%). You may also pass a survfit-compatible object directly.

**Typical use cases**

-   Draw one survival/CIF curve set by exposure groups (e.g., treatment vs control).
-   Call `cifpanel()` with a simplified code to create a panel displaying plots of multiple stratified survival/CIF curves or CIF curves for each event type.
-   Add CIs and censor/competing-risk/intercurrent-event marks.
-   Add a risk table to display the number at risk or the estimated survival probabilities or CIFs and CIs at each point in time.

**Key arguments shared with cifcurve()**

-   **Outcome type and estimator**
    -   `formula` A model formula specifying the outcome (`Event(time, status)` or `Surv(time, status)`) on the left-hand side (LHS) and, optionally, a stratification variable on the right-hand side (RHS). 
    -   `outcome.type = "survival"` → Kaplan–Meier estimator
    -   `outcome.type = "competing-risk"` → Aalen–Johansen estimator
-   **Confidence intervals**
    -   `conf.int` sets the two-sided level (default 0.95)\
    -   `conf.type` chooses the transformation (`"arcsine-square root"`, `"plain"`, `"log"`, `"log-log"`, `"logit"`, or `"none"`)
    -   `error` chooses the estimator for SE (`"greenwood"`, `"tsiatis"` or `"if"` for survival curves and `"delta"`, `"aalen"` or `"if"` for CIFs)

**Key arguments for cifplot()**

-   **Data visualization**
    -   `add.conf` adds CIs on the ggplot2-based plot
    -   `add.competing.risk.mark` and `add.intercurrent.event.mark` add symbols to describe competing risks or intercurrent events in addition to conventional censoring marks with `add.censor.mark`
    -   `add.risktable` adds numbers at risk
    -   `add.estimate.table` adds time-by-time estimates and CIs
    -   `add.quantile` add a reference line at a chosen quantile level

-   **Plot customization**
    -   `type.y` chooses y-axis (`"surv"` for survival and `"risk"` for 1-survival/CIF)
    -   `limits.x`, `limits.y`, `breaks.x`, `breaks.y` numeric vectors for axis control
    -   `style` specifies the appearance of plot (`"classic"`, `"bold"`, `"framed"`, `"grid"`, `"gray"` or `"ggsurvfit"`)
    -   `palette` specifies color of each curve (e.g. `palette=c("blue1", "cyan3", "navy", "deepskyblue3"))`)

-   **Panel display**
    -   `panel.per.variable` produces multiple survival/CIF curves per stratification variable specified in the formula
    -   `panel.per.event` produces CIF curves for each event type
    -   `panel.censoring` produces the Kaplan–Meier curves for (event, censor) and (censor, event) so that censoring patterns can be inspected
    -   `panel.mode` uses automatic panel mode

**Return**

-   A **cifplot** object (list) with at least: 
    -   `plot` the **ggplot** object
    -   Additional metadata used by `cifpanel()` and helper functions

**Under the hood: cifcurve()**

`cifplot()` is a streamlined, opinionated wrapper around `cifcurve()`, which calculates the Kaplan–Meier estimator and the Aalen–Johansen estimator. `cifcurve()` returns a survfit-compatible object, enabling an independent use of standard methods such as:

-   `summary()` — time-by-time estimates with SEs and CIs
-   `plot()` — base R stepwise survival/CIF curves
-   `mean()` — restricted mean survival estimates with CIs
-   `quantile()` — quantile estimates with CIs

Furthermore, the cifmodeling package provides a workflow for a **ggplot2-based visualization** based on the survfit-compatible objects by passing them to `cifplot()` and `cifpanel`. 

### cifpanel()

`cifpanel()` arranges multiple `cifplot()`/ggplot (or `cifcurve()` results rendered via `cifplot()`) into a unified layout with consistent scales, shared legend, and synchronized themes. The inset feature allows you to display one plot inside another.

**Typical use cases**
-   Compare CIF (event 1) vs CIF (event 2) in a 1×2 layout.

-   Compare survival/CIF curves across strata with a shared legend and matched axes.

-   Display a plot with an enlarged y-axis inside a full-scale plot.

**Key arguments**

-   `formula` or `formulas` — one formula or a list of formulas; each entry creates a panel.
-   `data`, `outcome.type`, `code.events`, `type.y` — recycled across panels unless a list is supplied for per-panel control.
-   `rows.columns.panel` — selects grid layout by c(rows, cols).
-   `inset.panel` — selects inset layout.
-   `title.panel`, `subtitle.panel`, `caption.panel`, `title.plot` — overall titles and captions.
-   `tag.panel` — panel tag style (e.g., "A", "a", "1").
-   `label.x`, `label.y`, `limits.x`, `limits.y`, `breaks.x`, `breaks.y` — shared axis control unless a list is supplied for per-panel control. 

**Return**

-   A **cifpanel** object (list) with at least: 
    -   `patchwork` the **patchwork** object (still ggplot-compatible)
    -   `plots` a list of **ggplot** objects in the `patchwork`
    -   Additional metadata

**An example of usage**

In the first example, `cifplot()` called `cifpanel()` to combine multiple plots. `cifpanel()` is a function for creating a panel with multiple plots like this. `cifpanel()` can also display one plot inside another. The cumulative incidence probability for macrovascular complications is low, making it difficult to discern differences between groups. The following code uses `inset.panel=` to display plot `output2`, with an enlarged vertical axis, inside plot `output1`. The position and size of the zoomed-in-view panel is specified by `inset.left`, `inset.bottom`, `inset.right`, and `inset.top`. 

```{r example02-1, fig.cap="Zoomed-in-view panel using cifplot() and cifpanel()", warning=FALSE, message=FALSE}
output1 <- cifplot(Event(t,epsilon) ~ fruitq,
                   data=diabetes.complications,
                   outcome.type="competing-risk",
                   code.event1=2,
                   code.event2=1,
                   add.conf=FALSE,
                   add.risktable = FALSE,
                   label.y="CIF of macrovascular complications",
                   label.x="Years from registration")
output2 <- cifplot(Event(t,epsilon) ~ fruitq,
                   data=diabetes.complications,
                   outcome.type="competing-risk",
                   code.event1=2,
                   code.event2=1,
                   add.conf = FALSE,
                   add.risktable = FALSE,
                   label.y="",
                   label.x="",
                   limits.y=c(0,0.15))
output3 <- list(a = output1$plot, b = output2$plot)
output4 <- cifpanel(plots=output3,
         title.plot=c("Fruit intake and macrovascular complications", "Zoomed-in view"),
         inset.panel=TRUE,
         inset.left=0.40, inset.bottom=0.45,
         inset.right=1.00, inset.top=0.95,
         inset.legend.position="none",
         legend.position="bottom")
print(output4)
```

The code below, in which a formula and data is given directly, produces the same output as above. 

```{r example02-2, eval=FALSE, warning=FALSE, message=FALSE}
cifpanel(
 title.plot = c("Fruit intake and macrovascular complications", "Zoomed-in view"),
 inset.panel = TRUE,
 formula = Event(t, epsilon) ~ fruitq,
 data = diabetes.complications,
 outcome.type = "competing-risk",
 code.events = list(c(2,1,0), c(2,1,0)),
 label.y = c("CIF of macrovascular complications", ""),
 label.x = c("Years from registration", ""),
 limits.y = list(c(0,1), c(0,0.15)),
 inset.left = 0.40, inset.bottom = 0.45,
 inset.right = 1.00, inset.top = 0.95,
 inset.legend.position = "none",
 legend.position = "bottom", 
 add.conf = FALSE
)
```

## polyreg()

This function implements **log odds product modeling** of CIFs at user-specified time points, focusing on multiplicative effects of a categorical exposure, or constant effects over time like Cox regression and Fine-Gray models. It estimates multiplicative effects such as **risk ratios**, **odds ratios**, or **subdistribution hazard ratios**, while ensuring that the probabilities across competing events sum to one. This is achieved through **reparameterization using polytomous log odds products**, which fits so-called effect-measure models and nuisance models on multiple competing events simultaneously. Additionally, `polyreg()` supports direct binomial regression for a survival outcome and the Richardson model for a binomial outcome, both of which use log odds products.

The function follows the familiar **formula + data** syntax with `Event()` or `Surv()` and outputs tidy results, including point estimates, SEs, CIs, and p-values. Its results can be easily summarized with `summary()` or combined with tools such as **modelsummary** or **broom** for reporting.

**Key arguments**

-   `nuisance.model` — a formula describing the outcome and nuisance covariates, excluding the exposure of interest. 
-   `exposure` — specifies the exposure variable. This argument accepts one binary or categorical variable.
-   `effect.measure1` and `effect.measure2` — specifies the effect measures for event1 and event2 (`"RR"`, `"OR"` or `"SHR"`). 
-   `outcome.type` selects the outcome type (`"competing-risk"`, `"survival"`, `"binomial"`, `"proportional-survival"` or `"proportional-competing-risk"`). 
-   `time.point` — specifies time points at which the exposure effect is evaluated. Required for `"competing-risk"` and `"survival"` outcomes.
-   `strata` — specifies the stratification variable used for the IPCW adjustment for dependent censoring. 

**Return**

-   `coefficient` — regression coefficients

-   `cov` — variance-covariance matrix for regression coefficients

-   `diagnostic.statistics` — a data frame containing inverse probability weights, influence functions, and predicted potential outcomes

-   `summary` — a summary of estimated exposure effects

### An example of usage

```{r example02-3, eval=FALSE, warning=FALSE, message=FALSE}
output <- polyreg(nuisance.model=Event(t,epsilon)~1, exposure="fruitq", 
          data=diabetes.complications, effect.measure1="RR", 
          effect.measure2="RR", time.point=8, outcome.type="competing-risk")
```

## Example 1. Unadjusted competing risks analysis

For the initial illustration, unadjusted analysis focusing on cumulative incidence of diabetic retinopathy (event 1) and macrovascular complications (event 2) at 8 years of follow-up is demonstrated. To visualize each covariate separately when multiple strata are supplied, set `panel.per.variable = TRUE`. Each variable on the right-hand side is plotted in its own panel, and the layout can be controlled with `rows.columns.panel`. The figure below contrasts the CIFs of diabetic retinopathy for the quartiles `fruitq` and a binary exposure `fruitq1`, low (Q1) and high (Q2 to 4) intake of fruit, generated by `cifplot()`. The `add.conf=TRUE` argument adds CIs to the plot. This helps visualize the statistical uncertainty of estimated probabilities across exposure levels. When using numeric variables for stratification, discretize them beforehand with `cut()` or `factor()`. The labels of x-axis (Time) and y-axis (Cumulative incidence) in these panels are default labels. 

```{r example04-1-1, fig.cap="Cumulative incidence curves per each stratification variable", warning=FALSE, message=FALSE}
data(diabetes.complications)
diabetes.complications$fruitq1 <- ifelse(
  diabetes.complications$fruitq == "Q1","Q1","Q2 to Q4"
)
cifplot(Event(t,epsilon)~fruitq+fruitq1, data=diabetes.complications, 
        outcome.type="competing-risk",
        add.conf=TRUE, add.censor.mark=FALSE, 
        add.competing.risk.mark=FALSE, panel.per.variable=TRUE)
```

In the second figure, censoring marks are added along each curve (`add.censor.mark = TRUE`) to indicate individuals who were censored before experiencing any event. These marks visualize the timing and frequency of censoring, allowing a clearer understanding of loss-to-censoring patterns over follow-up. Here the workflow differs slightly from the previous code. First, we compute a survfit-compatible object `output1` using `cifcurve()` with `outcome.type="competing-risk"` by calculating the Aalen–Johansen estimator stratified by `fruitq1`. Then, `cifplot()` is used to generate the figure. The `label.y`, `label.x` and `limit.x` arguments are also used to customize the axis labels and limits. 

```{r example04-1-2, fig.cap="Cumulative incidence curves with censor marks and axis labels", warning=FALSE, message=FALSE}
output1 <- cifcurve(Event(t,epsilon)~fruitq1, data=diabetes.complications, 
                    outcome.type="competing-risk")
cifplot(output1, add.conf=FALSE, add.risktable=FALSE, 
        add.censor.mark=TRUE, add.competing.risk.mark=FALSE, 
        label.y="CIF of diabetic retinopathy", label.x="Years from registration",
        limits.x=c(0,8))
```

Next, competing risk marks are added (`add.competing.risk.mark = TRUE`) to indicate individuals who experienced the competing event (macrovascular complications) before diabetic retinopathy. The time points at which the macrovascular complications occurred were obtained as `output2` for each strata using a helper function `extract_time_to_event()`. These symbols help distinguish between events due to the primary cause and those attributable to competing causes. Note that the names of `competing.risk.time` and `intercurrent.event.time` must match the strata labels used in the plot if supplied by the user.

```{r example04-1-3, fig.cap="Cumulative incidence curves with competing risk marks", warning=FALSE, message=FALSE}
output2 <- extract_time_to_event(Event(t,epsilon)~fruitq1, 
                                 data=diabetes.complications, which_event="event2")
cifplot(output1, add.conf=FALSE, add.risktable=FALSE, 
        add.censor.mark=FALSE, add.competing.risk.mark=TRUE, competing.risk.time=output2, 
        label.y="CIF of diabetic retinopathy", label.x="Years from registration",
        limits.x=c(0,8))
```

The `label.strata` is another argument for customizing labels, but when inputting a survfit object, it becomes invalid because it does not contain stratum information. Therefore, the following code inputs the formula and data. `label.strata` is used by combining `level.strata` and `order.strata`. `level.strata` specifies the levels of the stratification variable corresponding to each label in `label.strata`. The levels specified in `level.strata` are then displayed in the figure in the order defined by `order.strata`. A figure enclosed in a square was generated, which is due to `style="framed"` specification.

```{r example04-1-4, fig.cap="Cumulative incidence curves with strata labels and FRAMED style", warning=FALSE, message=FALSE}
cifplot(Event(t,epsilon)~fruitq1, data=diabetes.complications, 
        outcome.type="competing-risk", add.conf=FALSE, add.risktable=FALSE, 
        add.estimate.table=TRUE, add.censor.mark=FALSE, add.competing.risk.mark=TRUE, 
        competing.risk.time=output2, label.y="CIF of diabetic retinopathy", 
        label.x="Years from registration", limits.x=c(0,8),
        label.strata=c("High intake","Low intake"), level.strata=c("Q2 to Q4","Q1"), 
        order.strata=c("Q1", "Q2 to Q4"), style="framed")
```

By specifying `add.estimate.table=TRUE`, the risks of diabetic retinopathy (estimates for CIFs) along with their CIs are shown in the table at the bottom of the figure The risk ratios at a specific time point (e.g. 8 years) for competing events can be jointly and coherently estimated using `polyreg()` with `outcome.type="competing-risk"`. In the code of `polyreg()` below, no covariates are included in the nuisance model (`~1` specifies intercept only). The effect of low fruit intake `fruitq1` is estimated as an unadjusted risk ratio (`effect.measure1="RR"`) for diabetic retinopathy (event 1) and macrovascular complications (event 2) at 8 years (`time.point=8`). 

```{r example04-1-5, fig.cap="Joint estimation of unadjusted risk ratios at 8 years using polyreg()", warning=FALSE, message=FALSE}
output3 <- polyreg(nuisance.model=Event(t,epsilon)~1, exposure="fruitq1", 
          data=diabetes.complications, effect.measure1="RR", effect.measure2="RR", 
          time.point=8, outcome.type="competing-risk", 
          report.nuisance.parameter=TRUE)
print(output3$coefficient)
print(output3$cov)
```

The summaries of analysis results in the list of outputs (e.g. `output3$summary` below) are compatible with the `modelsummary` ecosystem, which makes it easy to create publication-ready tables in a variety of formats. A typical workflow is to pass the `summary` element of a `polyreg()` fit (or a list of such fits) to `modelsummary::msummary()`. In this case, all regression coefficients are included in summary by setting `report.nuisance.parameter = TRUE`. Model summaries can also be exponentiated to display risk ratios, odds ratios, or subdistribution hazard ratios using `exponentiate` option. The summaries can be displayed in the Viewer with customized statistics such as p-values or CIs.

## Example 2. Survival analysis

The second example is time to first event analysis (`outcome.type="survival"`) to estimate the effect on the risk of diabetic retinopathy or macrovascular complications at 8 years. In the code below, `cifplot()` is directly used to generate a survfit-compatible object internally and plot it. 

```{r example04-2-1, fig.cap="Survival curves from cifplot()", warning=FALSE, message=FALSE}
diabetes.complications$d <- as.integer(diabetes.complications$epsilon>0)
cifplot(Event(t,d) ~ fruitq1, data=diabetes.complications, 
outcome.type="survival", add.conf=TRUE, add.censor.mark=FALSE, 
add.competing.risk.mark=FALSE, label.y="Survival probability", 
label.x="Years from registration", label.strata=c("High intake","Low intake"),
level.strata=c("Q2 to Q4","Q1"), order.strata=c("Q1", "Q2 to Q4"))
```

The code below specifies the Richardson model on the risk of diabetic retinopathy or macrovascular complications at 8 years (outcome.type="survival"). Dependent censoring is adjusted by stratified IPCW method (`strata='strata'`). Estimates other than the effects of exposure (e.g. intercept) are suppressed when `report.nuisance.parameter` is not specified.

```{r example04-2-2, fig.cap="Cause-specific estimation of an unadjusted risk ratio at 8 years using polyreg()", warning=FALSE, message=FALSE}
output4 <- polyreg(nuisance.model=Event(t,d)~1, 
          exposure="fruitq1", strata="strata", data=diabetes.complications,
          effect.measure1="RR", time.point=8, outcome.type="survival")
```
