---
title: "Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Load required packages for the vignette
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/vignette-",
  fig.width  = 6,   
  fig.height = 4,   
  dpi        = 300,     
  fig.retina = 2,       
  out.width  = "70%",   
#  dev = "ragg_png",     
  comment = "#>"
)

.has_cifmodeling <- requireNamespace("cifmodeling", quietly = TRUE)
if (.has_cifmodeling) {
library(cifmodeling)
}
set.seed(1)
```

<!-- badges: start --> <!-- badges: end -->

```{r setup, include=FALSE, warning=FALSE}
#pak::pak("gestimation/cifmodeling")
#devtools::install_github("gestimation/cifmodeling") 
#library(cifmodeling)
library(nleqslv)
library(boot)
library(ggplot2)
library(ggsurvfit)
library(patchwork)
```

## Example 1. Unadjusted competing risks analysis

For the initial illustration, unadjusted analysis focusing on cumulative incidence of diabetic retinopathy (event 1) and macrovascular complications (event 2) at 8 years of follow-up is demonstrated. To visualize each covariate separately when multiple strata are supplied, set `printEachVar = TRUE`. Each variable on the right-hand side is plotted in its own panel, and the layout can be controlled with `rows.columns.panel`. The figure below contrasts the cumulative incidence curves of diabetic retinopathy for quartile `fruitq1` and a binary exposure `fruitq1`, low (Q1) and high (Q2 to 4) intake of fruit, generated by `cifplot()`. The `addConfidenceInterval=TRUE` argument adds confidence intervals to the plot. This helps visualize the statistical uncertainty of estimated probabilities across exposure levels. When using numeric variables for stratification, discretize them beforehand with `cut()` or `factor()`. The labels of x-axis (Time) and y-axis (Cumulative incidence) in these panels are default labels. 

```{r example1-1, warning=FALSE, message=FALSE}
data(diabetes.complications)
diabetes.complications$fruitq1 <- ifelse(
  diabetes.complications$fruitq == "Q1","Q1","Q2 to Q4"
)
cifplot(Event(t,epsilon)~fruitq+fruitq1, data=diabetes.complications, 
        outcome.type="COMPETING-RISK",
        addConfidenceInterval=TRUE, addCensorMark=FALSE, 
        addCompetingRiskMark=FALSE, printEachVar = TRUE)
```

In the second graph, censoring marks are added along each curve (`addCensorMark = TRUE`) to indicate individuals who were censored before experiencing any event. These marks visualize the timing and frequency of censoring, allowing a clearer understanding of loss-to-censoring patterns over follow-up. Here the workflow differs slightly from the previous code. First, we compute a survfit-compatible object `output1` using `cifcurve()` with `outcome.type="COMPETING-RISK"` by calculating Aalen–Johansen estimator stratified by fruitq1. Then, `cifplot()` is used to generate the figure. The `label.x` and `label.y` arguments are also used to customize the axis labels. 


```{r example1-2, warning=FALSE, message=FALSE}
output1 <- cifcurve(Event(t,epsilon)~fruitq1, data=diabetes.complications, 
                    outcome.type="COMPETING-RISK")
cifplot(output1, addConfidenceInterval=FALSE, addRiskTable=FALSE, 
        addCensorMark=TRUE, addCompetingRiskMark=FALSE, 
        label.y="CIF of diabetic retinopathy", label.x="Years from registration")
```

Next, competing risk marks are added (`addCompetingRiskMark = TRUE`) to indicate individuals who experienced the competing event (macrovascular complications) before diabetic retinopathy. The time points at which the macrovascular complications occurred were obtained as `output2` for each strata using a helper function `extract_time_to_event()`. These symbols help distinguish between events due to the primary cause and those attributable to competing causes. Note that the names of `competing.risk.time` and `intercurrent.event.time` must match the strata labels used in the plot if supplied by the user.

```{r example1-3, warning=FALSE, message=FALSE}
output2 <- extract_time_to_event(Event(t,epsilon)~fruitq1, 
                                 data=diabetes.complications, which_event="event2")
cifplot(output1, addConfidenceInterval=FALSE, addRiskTable=FALSE, 
        addCensorMark=FALSE, addCompetingRiskMark=TRUE, competing.risk.time=output2, 
        label.y="CIF of diabetic retinopathy", label.x="Years from registration")
```

`label.strata` is an argument for customizing labels, but when inputting a survfit object, it becomes invalid because it does not contain stratum information.
Therefore, the following code inputs the formula and data. `label.strata` is used by combining `level.strata` and `order.strata`. `level.strata` specifies the levels of the stratification variable corresponding to each label in `label.strata`. The levels specified in `level.strata` are then displayed in the graph in the order defined by `order.strata`. A graph enclosed in a square was generated, which is due to `style="FRAMED"` specification.

```{r example1-4, warning=FALSE, message=FALSE}
cifplot(Event(t,epsilon)~fruitq1, data=diabetes.complications, 
        outcome.type="COMPETING-RISK", addConfidenceInterval=FALSE, addRiskTable=FALSE, 
        addEstimateTable=TRUE, addCensorMark=FALSE, addCompetingRiskMark=TRUE, 
        competing.risk.time=output2, label.y="CIF of diabetic retinopathy", 
        label.x="Years from registration", label.strata=c("High intake","Low intake"), 
        level.strata=c("Q2 to Q4","Q1"), order.strata=c("Q1", "Q2 to Q4"), style="FRAMED")
```

By specifying `addEstimateTable=TRUE`, the risks of diabetic retinopathy (estimates for CIFs) along with their confidence interval are shown in the table at the bottom of the graph. The risk ratios at a specific time point for competing events can be jointly and coherently estimated using `polyreg()` with `outcome.type="COMPETING-RISK"`. In the code of `polyreg()` below, no covariates are included in the nuisance model (`~1` specifies intercept only). The effect of low fruit intake `fruitq1` is estimated as an unadjusted risk ratio (`effect.measure1="RR"`) for diabetic retinopathy (event 1) and macrovascular complications (event 2) at 8 years (`time.point=8`). 

```{r example1-5, warning=FALSE, message=FALSE}
output3 <- polyreg(nuisance.model=Event(t,epsilon)~1, exposure="fruitq1", 
          data=diabetes.complications, effect.measure1="RR", effect.measure2="RR", 
          time.point=8, outcome.type="COMPETING-RISK", 
          report.nuisance.parameter=TRUE)
print(output3$coefficient)
print(output3$cov)
```

The summaries of analysis results in the list of outputs (e.g. `output3$summary` below) are compatible with the `modelsummary` ecosystem, which makes it easy to create publication-ready tables in a variety of formats. A typical workflow is to pass the `summary` element of a `polyreg()` fit (or a list of such fits) to `modelsummary::msummary()`. In this case, all regression coefficients are included in summary by setting `report.nuisance.parameter = TRUE`. Model summaries can also be exponentiated to display risk ratios, odds ratios, or subdistribution hazard ratios using `exponentiate` option. The summaries can be displayed in the Viewer with customized statistics such as p-values or confidence intervals.

## Example 2. Survival analysis

The second example is time to first event analysis (`outcome.type="SURVIVAL"`) to estimate the effect on the risk of diabetic retinopathy or macrovascular complications at 8 years. In the code below, `cifplot()` is directly used to generate a survfit-compatible object internally and plot it. 

```{r example2-1, warning=FALSE, message=FALSE}
data(diabetes.complications)
diabetes.complications$d <- as.integer(diabetes.complications$epsilon>0)
cifplot(Event(t,d) ~ fruitq1, data=diabetes.complications, 
outcome.type="SURVIVAL", addConfidenceInterval=TRUE, addCensorMark=FALSE, 
addCompetingRiskMark=FALSE, label.y="Survival probability", 
label.x="Years from registration", label.strata=c("High intake","Low intake"))
```

The code below specifies the Richardson model on the risk of diabetic retinopathy or macrovascular complications at 8 years (outcome.type="SURVIVAL"). Dependent censoring is adjusted by stratified IPCW method (`strata='strata'`). Estimates other than the effects of exposure (e.g. intercept) are suppressed when `report.nuisance.parameter` is not specified.

```{r example2-2, warning=FALSE, message=FALSE}
output4 <- polyreg(nuisance.model=Event(t,d)~1, 
          exposure="fruitq1", strata="strata", data=diabetes.complications,
          effect.measure1="RR", time.point=8, outcome.type="SURVIVAL")
```

## Example 3. Adjusted competing risks analysis

The code below specifies polytomous log-odds models of both of competing events (`outcome.type="COMPETING-RISK"`). Here 15 covariates and censoring strata are specified in `nuisance.model=` and `strata=`, respectively.

```{r example3, warning=FALSE, message=FALSE}
output5 <- polyreg(nuisance.model=Event(t,epsilon)~age+sex+bmi+hba1c
          +diabetes_duration+drug_oha+drug_insulin+sbp+ldl+hdl+tg
          +current_smoker+alcohol_drinker+ltpa, 
          exposure="fruitq1", strata="strata", data=diabetes.complications,
          effect.measure1="RR", time.point=8, outcome.type="COMPETING-RISK")
```

## Example 4. Description of cumulative incidence of competing events

`cifpanel()` arranges multiple survival and CIF plots into a single, polished layout with a shared legend. It’s designed for side-by-side comparisons—e.g., event 1 vs event 2, different groupings, or different y-scales—while keeping axis ranges and styles consistent. Internally each panel is produced using the same engine as `cifcurve()`, and you can supply scalar arguments (applied to all panels) or lists to control each panel independently.

This function accepts both shared and panel-specific arguments. When a single formula is provided, the same model structure is reused for each panel, and arguments supplied as lists are applied individually to each panel. Arguments such as `code.events`, `label.strata`, or `addCensorMark` can be given as lists, where each list element corresponds to one panel. This allows flexible configuration while maintaining a concise and readable syntax.

The example below creates a 1×2 panel (`rows.columns.panel = c(1,2)`) comparing the cumulative incidence of two competing events in the same cohort, namely CIF of diabetic retinopathy in the left panel and CIF of macrovascular complications in the right panel. Both panels are stratified by `fruitq1`, and the legend is shared at the bottom. The pairs of `code.events` as a list instructs `cifpanel()` to display event 1 in the first panel and event 2 in the second panel, with event code 0 representing censoring.

```{r example4, warning=FALSE, message=FALSE}
cifpanel(
 title.panel = "A comparison of cumulative incidence of competing events",
 rows.columns.panel = c(1,2),
 formula = Event(t, epsilon) ~ fruitq1,
 data = diabetes.complications,
 outcome.type = "COMPETING-RISK",
 code.events = list(c(1,2,0), c(2,1,0)),
 label.y = c("CIF of diabetic retinopathy", "CIF of macrovascular complications"),
 label.x = "Years from registration",
 label.strata = list(c("High intake","Low intake")),
 title.plot = c("Diabetic retinopathy", "Macrovascular complications"),
 legend.position = "bottom",
 legend.collect=TRUE
)
```

Arguments specified as scalars (for example, `label.x = "Years from registration"`) are applied uniformly to all panels. Character vectors of the same length as the number of panels (for example, `label.y = c("Diabetic retinopathy", "Macrovascular complications")`) assign a different label to each panel in order. Lists provide the most flexibility, allowing each panel to have distinct settings that mirror the arguments of `cifcurve()`.

The `legend.collect = TRUE` option merges legends from all panels into a single shared legend, positioned according to `legend.position`. The arguments `title.panel`, `subtitle.panel`, `caption.panel`, and `title.plot` control the overall panel title and individual subplot titles, ensuring that multi-panel layouts remain consistent and publication-ready. 

## Additional information

### Comparison of model formula and outcome.type across cifcurve() and polyreg()

**Common ground**

-   Both functions expect a `Surv()` or `Event()` response on the left-hand side of the model formula so that follow-up time and event indicators are supplied in the same way. Internally the helpers validate non-negative times, check that the event codes match the user-specified `code.event*` arguments, and derive indicators used downstream.
-   The `outcome.type` argument is normalized by the same utility, meaning that the same set of synonyms (e.g. `"C"`, `"competing risk"`, or `"S"`) are accepted in both workflows and mapped onto the canonical labels used in the estimators.

**Key differences**

-   In `cifcurve()` the right-hand side of the formula defines the grouping variable(s) whose curves are estimated; weights (if any) are supplied separately. There is no dedicated exposure argument because all strata are handled through the formula term that is turned into a factor for plotting and estimation.
-   In `polyreg()` the `nuisance.model` formula describes only the outcome and nuisance covariates (i.e. confounders), while the exposure is passed separately through the `exposure` argument. The exposure design matrix, reference level handling, and scaling of nuisance covariates are handled explicitly for regression fitting.
-   `cifcurve()` focuses on two estimands, selecting between Kaplan–Meier and Aalen–Johansen calculations via `outcome.type = "SURVIVAL"` or `"COMPETING-RISK"`. In contrast, `polyreg()` supports additional time-constant and binomial estimands (`"PROPORTIONAL-SURVIVAL"`, `"PROPORTIONAL-COMPETING-RISK"`, `"BINOMIAL"`) and enforces time-point requirements accordingly—for example `time.point` is mandatory for survival/competing-risk contrasts, while time-constant-effect (e.g. proportional subdistribution hazards) models can either use all observed event times or a user-specified grid supplied via a (vector) `time.point`.

### Event coding conventions in package cifmodeling and CDISC ADaM ADTTE

When working with survival data, one common source of confusion arises from inconsistent event coding conventions between the standard input of `Surv()`, provided by survival package, and the CDISC ADaM ADTTE (Analysis Data Model for Time-to-Event Endpoints) used in regulatory clinical trials. With `Surv()`, event is often coded as 1 and censoring is coded as 0. By contrast, the CDISC ADaM ADTTE (commonly used in pharmaceutical submissions and SDTM-based analyses) follows the opposite convention. `CNSR = 0` means that event occurred and `CNSR = 1` represents censoring, which aligns with the broader ADaM convention (where `1 = TRUE`). 

To accommodate differences in coding conventions, one can explicitly specify event and censoring codes using the `code.event1` and `code.censoring` arguments in `cifcurve()`, `cifplot()`, `cifpanel()` and `polyreg()`. This allows users to perform accurate survival analysis using functions from the cifmodeling package while maintaining the CDISC ADaM ADTTE coding scheme. 

```{r example5, eval=FALSE, warning=FALSE, message=FALSE}
polyreg(Event(AVAL, CNSR) ~ ARM, data = adtte, 
        outcome.type = "SURVIVAL", code.event1=0, code.censoring=1)
```

### Processing pipeline of cifcurve()

A user-specified model formula in `Event()` or `Surv()` is first parsed by `util_read_surv()` to extract the time-to-event, the event indicator, optional weights, and any strata variable. Those components are then passed to the back-end estimators implemented in `src/calculateAJ_Rcpp.cpp`. The C++ routine supports both weighted and stratified data, so the heavy lifting for Kaplan–Meier or Aalen–Johansen estimates (and associated at-risk, event, and censor counts) happens efficiently regardless of how the input was specified. 

- The `error` argument determines how standard errors are computed. For survival data (`outcome.type = "SURVIVAL"`) it maps to Greenwood (default) or Tsiatis estimators, while for competing-risks data (`outcome.type = "COMPETING-RISK"`) `"aalen"` reproduces the classic Aalen standard errors and `"delta"` implements the delta method.

### Variance and confidence interval options for cifcurve()

**Standard errors** For standard errors of Kaplan–Meier estimator, the default is Greenwood estimator, with alternatives `"tsiatis"` for the Tsiatis estimator The Greenwood formula is a sum of terms d/(n(n-m)), where d is the number of events at a given time point, n is the number at risk, and m is the cumulative sum of events at that time. This formula is derived from a binomial argument, so extension to the weighted case is ad hoc. Tsiatis (1981) proposes a sum of terms d/(n\*n), based on a counting process argument which includes the weighted case. For confidence intervals of cumulative incidence curves, the default `"delta"` applies the delta-method, while `"aalen"` uses the Aalen variance decomposition. By default `cifcurve()` rescales the Greenwood/Tsiatis quantities so that `std.err` is reported on the probability scale; set `report.survfit.std.err = TRUE` to return the conventional log-survival standard errors from `survfit()`.

**Confidence intervals** `src/calculateAJ_Rcpp.cpp` constructs intervals on the probability scale using the requested transformation: `"arcsine-square root"`/`"arcsine"`/`"a"` (default), `"plain`, `"log"`, `"log-log"`, or `"logit"`. Passing `"none"`/`"n"` skips interval computation entirely. The function exponentiates back to the probability scale, clips bounds to [0, 1], and replaces undefined values with `NA` so that interval endpoints remain well behaved in plots and summaries.

### Processing pipeline of polyreg()

Input pre-processing mirrors the survival workflow by calling `checkInput()` to parse event codes, exposure levels, and requested estimands before `createAnalysisDataset()` builds the modeling frame and `reg_normalize_covariate()` rescales covariates by default. Starting values are obtained from the nuisance model fits via `getInitialValues()` or `getInitialValuesProportional()`, or, when necessary, from user-supplied data frame of starting values. Inverse-probability-of-censoring weights for survival and competing-risks outcomes are calculated by `calculateIPCW()`, which in turn wraps the `calculateKM()` C++ routine to obtain Kaplan-Meier estimates of censoring, while time-constant effects assemble matrices through `calculateIPCWMatrix()`.

The weighted estimating equations are then solved by `solveEstimatingEquation()`, which dispatches to specific score functions such as `estimating_equation_ipcw()`, `estimating_equation_survival()`, and `estimating_equation_proportional()` and iterates with `nleqslv`/Levenberg–Marquardt updates until the convergence criteria in `assessConvergence()` are met. Sandwich variances come from `calculateCov()` or `calculateCovSurvival()` and are rescaled back to the original covariate scale by `reg_normalize_estimate()`, while optional bootstrap intervals reuse the same solver inside the `boot` resampling loop. The resulting influence functions, weights, and fitted cumulative incidence probabilities are retained in `diagnostic.statistics` for downstream diagnostics.

### Reproducibility and conventions for polyreg()

 - If convergence warnings appear, relax/tighten tolerances or cap the parameter bound (`optim.parameter1` to `3`) and inspect `report.optim.convergence = TRUE`.
 - If necessary, try modifying other `optim.parameter`, providing user-specified initial values, or reducing the number of nuisance parameters (e.g. provide a `time.point` grid that contributes to estimation when using `"PROPORTIONAL-SURVIVAL"` and `"PROPORTIONAL-COMPETING-RISK"`).
 - Set `boot.seed` for reproducible bootstrap results.
 - Match CDISC ADaM conventions via `code.event1 = 0`, `code.censoring = 1` (and, if applicable, `code.event2` for competing events).

### Key arguments of msummary() that are helpful when reporting polyreg() results

-   `output` – controls the destination of the table (e.g., `"markdown"`, `"latex"`, `"html"`, or a file path such as `'results/cif_table.docx'`).
-   `coef_map` or `coef_rename` – renames model terms for interpretability.
-   `statistic` – which uncertainty summaries to display. Examples: `statistic = "p.value"` for p-values, `statistic = "conf.int"` for Wald confidence interval, `statistic = "t"` for t statistics, `and statistic = "z"` for z statistics, instead of standard errors (default).

You can also supply multiple model summaries as a named list to compare estimates across specifications in a single table. See `vignette('modelsummary')` for a comprehensive overview of additional layout and styling options.
