---
title: "Package cifmodeling: getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package cifmodeling: getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Load required packages for the vignette
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/vignette-",
  out.width = "100%"
)

if (!requireNamespace("cifmodeling", quietly = TRUE)) {
  stop("The 'cifmodeling' package must be installed to run this vignette.")
}

knitr::opts_chunk$set(
  fig.width  = 6,   
  fig.height = 4,   
  dpi        = 300,     
  fig.retina = 2,       
  out.width  = "70%",   
#  dev = "ragg_png",     
  comment = "#>"
)
```

<!-- badges: start --> <!-- badges: end -->

```{r setup, include=FALSE, warning=FALSE}
#pak::pak("gestimation/cifmodeling")
#devtools::install_github("gestimation/cifmodeling") 
library(cifmodeling)
library(nleqslv)
library(boot)
library(ggplot2)
library(ggsurvfit)
library(patchwork)
```


## Quick start

```{r syntax1, warning=FALSE, message=FALSE}
data(diabetes.complications)
cifplot(Event(t,epsilon) ~ fruitq, data=diabetes.complications, 
        outcome.type="COMPETING-RISK", printEachEvent=TRUE)
```

This is an example code snippet applying `cifplot()` to create cumulative incidence function (CIF) curves. In competing risks data, censoring is often coded as 0, events of interest as 1, and competing risks as 2. The variable `epsilon` in `diabetes.complications` data frame represents the occurrence of competing risks according to this coding scheme. By specifying `printEachEvent=`, the CIF curve for diabetic retinopathy (`epsilon=1`) is output on the left, and the CIF curve for macrovascular complications (`epsilon=2`) is output on the right. 

## Overview

In medical and epidemiological research, analysts often need to handle censoring, competing risks, and intercurrent events (e.g., treatment switching), but existing R packages typically separate these tasks into different interfaces. `cifmodeling` provides a **unified, publication-ready toolkit** that integrates description of survival and cumulative incidence function (CIF) curves, regression modeling, and visualization for survival and competing risks data. It covers both nonparametric estimation and regression modeling of CIFs, centered around three tightly connected functions.

-   `cifplot()` typically generates a survival or cumulative incidence curve with marks that represent censoring, competing risks and intercurrent events. Multiple variance estimators and confidence interval methods are supported. Visualization relies on `ggsurvfit/ggplot2`. 

-   `cifpanel()` generates a multi-panel figure for survival/CIF curves, arranged either in a grid layout or as an inset overlay.

-   `polyreg()` fits coherent regression models of CIFs based on polytomous log odds products and the stratified IPCW estimator. This function is particularly well-suited for causal inference in terms of typical effect measures, namely risk ratios, odds ratios, and subdistribution hazard ratios, with a competing risks, survival, or binary outcome. 

These functions adopt a formula + data syntax, return tidy, publication-ready outputs, and integrate seamlessly with `ggsurvfit` and `modelsummary` for visualization.

## Quality control

`cifmodeling` includes an extensive test suite built with **testthat**, ensuring the numerical accuracy and graphical consistency of all core functions (`cifcurve`, `cifplot`, `cifpanel`, and `polyreg`). The package is continuously tested on GitHub Actions (Windows, macOS, Linux) to maintain reproducibility and CRAN-level compliance.

## Installation

The package is implemented in R and relies on `Rcpp`, `nleqslv` and `boot` for the numerical back-end. The examples in this README also use `ggplot2`, `ggsurvfit`, `patchwork` and `modelsummary` for tabulation and plotting. Install the core package and these companion packages with:

``` r
install.packages("Rcpp")
install.packages("nleqslv")
install.packages("boot")
install.packages("ggplot2")
install.packages("ggsurvfit")
install.packages("patchwork")
install.packages("modelsummary")
devtools::install_github("gestimation/cifmodeling")
```

## Main functions

### cifplot()

This function produces the **Kaplan–Meier survival** or **Aalen–Johansen cumulative incidence** curve from a unified formula + data interface. It auto-labels axes based on `outcome.type` and `type.y`, can add censoring/competing-risk/intercurrent-event marks, and returns a regular ggplot object (compatible with + and %+%). You may also pass a survfit-compatible object directly.

**Typical use cases**

-   Draw one survival/CIF curve set by exposure groups (e.g., treatment vs control).
-   Call `cifpanel()` with a simplified code to create a panel displaying plots of multiple stratified survival/CIF curves or CIF curves for each event type.
-   Add confidence intervals and censor/competing-risk/intercurrent-event marks.
-   Add a risk table to display the number at risk or the estimated survival probabilities or CIFs and 95% confidence intervals at each point in time.

**Key arguments shared with cifcurve()**

-   **Outcome type and estimator**
    -   `formula` A model formula specifying the time-to-event outcome on the left-hand side (`Event(time, status)` or `survfit::Surv(time, status)`) and, optionally, a stratification variable on the right-hand side. 
    -   `outcome.type = "SURVIVAL"` → Kaplan–Meier estimator
    -   `outcome.type = "COMPETING-RISK"` → Aalen–Johansen estimator
-   **Confidence intervals**
    -   `conf.int` sets the two-sided level (default 0.95)\
    -   `conf.type` chooses the transformation (`"arcsin"`, `"plain"`, `"log"`, `"log-log"`, `"logit"`, or `"none"`)
    -   `error` chooses the estimator for standard error (`"greenwood"` or `"tsiatis"` for survival curves and `"delta"` or `"aalen"` for CIFs)

**Key arguments for cifplot()**

-   **Data visualization**
    -   `addConfidenceInterval` adds confidence intervals on the ggplot2-based plot
    -   `addCompetingRiskMark` and `addIntercurrentEventMark` add symbols to describe competing risks or intercurrent events in addition to conventional censoring marks with `addCensorMark`
    -   `addRiskTable` adds numbers at risk
    -   `addEstimateTable` adds estimates and 95% confidence interval
    -   `addQuantileLine` adds a line that represents median or quantile

-   **Panel display**
    -   `printEachVar` produces multiple survival/CIF curves per stratification variable specified in the formula
    -   `printEachEvent` produces CIF curves for each event type
    -   `printCensoring` produces KM-type curves for (event, censor) and (censor, event) so that censoring patterns can be inspected
    -   `panel.mode` uses automatic panel mode

-   **Plot customization**
    -   `type.y` chooses y-axis. (`"surv"` for survival curves and `"risk"` for CIFs)
    -   `limits.x`, `limits.y`, `breaks.x`, `breaks.y` numeric vectors for axis control
    -   `style` specifies the appearance of plot (`"CLASSIC"`, `"BOLD"`, `"FRAMED"`, `"GRID"`, `"GRAY"` or `"GGSURVFIT"`)

**Return**

-   A **ggplot** object.

**Under the hood: cifcurve()**

`cifplot()` is a streamlined, opinionated wrapper around `cifcurve()`, which calculates the Kaplan–Meier estimator and the Aalen–Johansen estimator. `cifcurve()` returns a survfit-compatible object, enabling an independent use of standard methods such as:

-   `summary()` — time-by-time estimates with standard errors and confidence intervals
-   `plot()` — base R stepwise survival/CIF curves
-   `mean()` — restricted mean survival estimates with confidence intervals
-   `quantile()` — quantile estimates with confidence intervals

Furthermore, the cifmodeling package provides a workflow for a **ggplot2-based visualization** based on the survfit-compatible objects by passing them to `cifplot()` and `cifpanel`. 

### cifpanel()

`cifpanel()` arranges multiple `cifplot()/cifcurve()` results into a unified grid layout with consistent scales, shared legend, and synchronized themes. It is ideal for visual comparison across outcomes, time scales, or strata. The inset feature also allows you to display another plot within a plot.

**Typical use cases**
-   Compare CIF (event 1) vs CIF (event 2) in a 1×2 layout.

-   Compare survival/CIF curves across strata with a shared legend and matched axes.

-   Display a plot with an enlarged y-axis within a plot scaled from 0 to 100%. 

**Key arguments**

-   `formula` or `formulas` — one formula or a list of formulas; each entry creates a panel.
-   `data`, `outcome.type`, `code.events`, `type.y` — recycled across panels unless a list is supplied for per-panel control.
-   `rows.columns.panel` — selects grid layout by c(rows, cols).
-   `inset.panel` — selects inset layout.
-   `title.panel`, `subtitle.panel`, `caption.panel`, `title.plot` — overall titles and captions.
-   `tag.panel` — panel tag style (e.g., "A", "a", "1").
-   `label.x`, `label.y`, `limits.x`, `limits.y`, `breaks.x`, `breaks.y` — shared axis control unless a list is supplied for per-panel control. 

**Return**

-   A **patchwork** object (still ggplot-compatible).

**An example of usage**

In the first example, `cifplot()` called `cifpanel()` to combine multiple plots. `cifpanel()` is a function for creating a panel with multiple plots like this. `cifpanel()` can also display one plot inside another. The cumulative incidence probability for macrovascular complications is low, making it difficult to discern differences between groups. The following code uses `inset.panel=` to display plot `output2`, with an enlarged vertical axis, inside plot `output1`. The position and size of the zoomed-in-view panel is specified by `inset.left`, `inset.bottom`, `inset.right`, and `inset.top`. 

```{r syntax2, warning=FALSE, message=FALSE}
output1 <- cifplot(Event(t,epsilon) ~ fruitq,
                   data = diabetes.complications,
                   outcome.type="COMPETING-RISK",
                   code.event1=2,
                   code.event2=1,
                   addConfidenceInterval = FALSE,
                   addRiskTable = FALSE,
                   label.y="CIF of macrovascular complications",
                   label.x="Years from registration")
output2 <- cifplot(Event(t,epsilon) ~ fruitq,
                   data = diabetes.complications,
                   outcome.type="COMPETING-RISK",
                   code.event1=2,
                   code.event2=1,
                   addConfidenceInterval = FALSE,
                   addRiskTable = FALSE,
                   label.y="",
                   label.x="",
                   limits.y=c(0,0.15))
output3 <- list(a=output1, b=output2)
cifpanel(plots = output3,
         title.plot = c("Associations between fruit intake and macrovascular complications", "Zoomed-in view"),
         inset.panel = TRUE,
         inset.left = 0.40, inset.bottom = 0.45,
         inset.right = 1.00, inset.top = 0.95,
         inset.legend.position = "none",
         legend.position = "bottom")
```

The code below, in which a formula and data is given directly, produces the same output as above. 

```{r syntax3, eval=FALSE, warning=FALSE, message=FALSE}
data(diabetes.complications)
cifpanel(
 title.plot = c("Associations between fruit intake and macrovascular complications", "Zoomed-in view"),
 inset.panel = TRUE,
 formula = Event(t, epsilon) ~ fruitq,
 data = diabetes.complications,
 outcome.type = "COMPETING-RISK",
 code.events = list(c(2,1,0), c(2,1,0)),
 label.y = c("CIF of macrovascular complications", ""),
 label.x = c("Years from registration", ""),
 limits.y = list(c(0,1), c(0,0.15)),
 inset.left = 0.40, inset.bottom = 0.45,
 inset.right = 1.00, inset.top = 0.95,
 inset.legend.position = "none",
 legend.position = "bottom", 
 addConfidenceInterval = FALSE
)
```

### polyreg()

`polyreg()` implements **log odds product modeling** for CIFs at user-specified time points, focusing on multiplicative effects of a categorical exposure, or constant effects over time like Cox regression and Fine-Gray models. It estimates multiplicative effects such as **risk ratios**, **odds ratios**, or **subdistribution hazard ratios**, while ensuring that the probabilities across competing events sum to one. This is achieved through **reparameterization using polytomous log odds products**, which fits so-called effect-measure models and nuisance models on multiple competing events simultaneously. Additionally, `polyreg()` supports direct binomial regression for survival outcomes and the Richardson model for binomial outcomes, both of which use log odds products.

The function follows the familiar **formula + data** syntax with `Event()` or `Surv()` and outputs tidy results, including point estimates, standard errors, confidence intervals, and p-values. Its results can be easily summarized with `summary()` or combined with tools such as **modelsummary** or **broom** for reporting.

**Key arguments**

-   `nuisance.model` — a formula describing the outcome and nuisance covariates, excluding the exposure of interest. 
-   `exposure` — specifies the exposure variable. This argument accepts one binary or categorical variable.
-   `effect.measure1` and `effect.measure2` — specifies the effect measures for event1 and event2 (`"RR"`, `"OR"` or `"SHR"`). 
-   `outcome.type` selects the outcome type (`"COMPETING-RISK"`, `"SURVIVAL"`, `"BINOMIAL"`, `"PROPORTIONAL-SURVIVAL"` or `"PROPORTIONAL-COMPETING-RISK"`). 
-   `time.point` — specifies time points at which the exposure effect is evaluated. Required for `"COMPETING-RISK"` and `"SURVIVAL"` outcomes.
-   `strata` — specifies the stratification variable used to adjust for dependent censoring. 

**Return**

-   `coefficient` — regression coefficients

-   `cov` — variance-covariance matrix for regression coefficients

-   `diagnostic.statistics` — a data frame containing inverse probability weights, influence functions, and predicted potential outcomes

-   `summary` — a summary of estimated exposure effects

### An example of usage

```{r syntax4, eval=FALSE, warning=FALSE, message=FALSE}
data(diabetes.complications)
output <- polyreg(nuisance.model=Event(t,epsilon) ~ +1, exposure="fruitq", 
          data=diabetes.complications, effect.measure1="RR", 
          effect.measure2="RR", time.point=8, outcome.type="COMPETING-RISK")
```

If you only need the core functionality of `cifcurve()` and `polyreg()`, installing `Rcpp` and `nleqslv` are enough; the other packages are optional but recommended to reproduce the examples below.

## Direct polytomous modeling of CIFs

The model for `polyreg()` is specified by three main components:

-   Nuisance model: Describes the relationship between outcomes and covariates (excluding exposure).

-   Effect measures and time points: Defines the exposure effect to be estimated and the time point of interest.

-   Censoring adjustment: Specifies strata for inverse probability weighting to adjust for dependent censoring.

**1. Nuisance Model**

The `nuisance.model` argument specifies the formula linking the outcome to covariates. Its format depends on the outcome type:

-   Competing risks or survival outcome: Use `Surv()` or `Event()` with time and status variables.

-   Binomial outcome: Use standard R formula notation.

Default event codes:

-   Competing risks outcome: 1 and 2 for event types, 0 for censoring.

-   Survival outcome: 1 for events, 0 for censoring.

-   Binomial outcome: 0 and 1.

Event codes can be customized using `code.event1`, `code.event2`, and `code.censoring`. The `outcome.type` argument must be set to:

-   Effects on cumulative incidence probabilities at a specific time: `"COMPETING-RISK"`

-   Effects on a risk at a specific time: `"SURVIVAL"`

-   Common effects on cumulative incidence probabilities over time: `"PROPORTIONAL-COMPETING-RISK"`

-   Common effects on a risk over time: `"PROPORTIONAL-SURVIVAL"`

-   Effects on a risk of a binomial outcome: `"BINOMIAL"`

Covariates included in `nuisance.model` should adjust for confounding factors to obtain unbiased exposure effect estimates.

**2. Effect measures and time points**

Three effect measures available:

-   Risk ratio (RR)

-   Odds ratio (OR)

-   Subdistribution hazard ratio (SHR)

Set the desired measure using `effect.measure1` and, for competing risks analysis, `effect.measure2`. The `time.point` argument specifies the follow-up time at which effects are estimated.

**3. Censoring adjustment**

Inverse probability weights adjust for dependent censoring. Use `strata=` to specify stratification variables. If no strata are specified, Kaplan-Meier weights are used.

**Return**

This function returns a list object that includes:

-   `coefficient` — regression coefficients

-   `cov` — variance-covariance matrix for regression coefficients

-   `diagnostic.statistics` — a data frame containing inverse probability weights, influence functions, and predicted potential outcomes

-   `summary` — a summary of estimated exposure effects

Use `summary` output with `msummary()` to display formatted results. The regression coefficients and their variance-covariance matrix are provided as `coefficient` and `cov`, respectively, with the first element corresponding to the intercept term, subsequent elements to the covariates in `nuisance.model`, and the last element to the variable specified by `exposure=`. Finally, `diagnostic.statistics` is a data frame containing inverse probability weights, influence functions, and predicted values of the potential outcomes of individual observations. 

## Example 1. Unadjusted competing risks analysis

For the initial illustration, unadjusted analysis focusing on cumulative incidence of diabetic retinopathy (event 1) and macrovascular complications (event 2) at 8 years of follow-up is demonstrated. To visualize each covariate separately when multiple strata are supplied, set `printEachVar = TRUE`. Each variable on the right-hand side is plotted in its own panel, and the layout can be controlled with `rows.columns.panel`. The figure below contrasts the cumulative incidence curves of diabetic retinopathy for quartile `fruitq1` and a binary exposure `fruitq1`, low (Q1) and high (Q2 to 4) intake of fruit, generated by `cifplot()`. The `addConfidenceInterval=TRUE` argument adds confidence intervals to the plot. This helps visualize the statistical uncertainty of estimated probabilities across exposure levels. When using numeric variables for stratification, discretize them beforehand with `cut()` or `factor()`. The labels of x-axis (Time) and y-axis (Cumulative incidence) in these panels are default labels. 

```{r example1-1, warning=FALSE, message=FALSE}
data(diabetes.complications)
diabetes.complications$fruitq1 <- ifelse(
  diabetes.complications$fruitq == "Q1","Q1","Q2 to Q4"
)
cifplot(Event(t,epsilon)~fruitq+fruitq1, data=diabetes.complications, outcome.type="COMPETING-RISK",
        addConfidenceInterval=TRUE, addCensorMark=FALSE, addCompetingRiskMark=FALSE, printEachVar = TRUE)
```

In the second graph, censoring marks are added along each curve (`addCensorMark = TRUE`) to indicate individuals who were censored before experiencing any event. These marks visualize the timing and frequency of censoring, allowing a clearer understanding of loss-to-censoring patterns over follow-up. Here the workflow differs slightly from the previous code. First, we compute a survfit-compatible object `output1` using `cifcurve()` with `outcome.type="COMPETING-RISK"` by calculating Aalen–Johansen estimator stratified by fruitq1. Then, `cifplot()` is used to generate the figure. The `label.x` and `label.y` arguments are also used to customize the axis labels. 

```{r example1-2, warning=FALSE, message=FALSE}
output1 <- cifcurve(Event(t,epsilon)~fruitq1, data=diabetes.complications, outcome.type="COMPETING-RISK")
cifplot(output1, addConfidenceInterval=FALSE, addEstimateTable=TRUE, addCensorMark=TRUE, 
        addCompetingRiskMark=FALSE, label.y="CIF of diabetic retinopathy", label.x="Years from registration")
```

Next, competing risk marks are added (`addCompetingRiskMark = TRUE`) to indicate individuals who experienced the competing event (macrovascular complications) before diabetic retinopathy. The time points at which the macrovascular complications occurred were obtained as `output2` for each strata using a helper function `extract_time_to_event()`. These symbols help distinguish between events due to the primary cause and those attributable to competing causes. Note that the names of `competing.risk.time` and `intercurrent.event.time` must match the strata labels used in the plot if supplied by the user.


```{r example1-3, warning=FALSE, message=FALSE}
output2 <- extract_time_to_event(Event(t,epsilon)~fruitq1, 
                                 data=diabetes.complications, which_event="event2")
cifplot(output1, addConfidenceInterval=FALSE, addEstimateTable=TRUE, 
        addCensorMark=FALSE, addCompetingRiskMark=TRUE, competing.risk.time=output2, 
        label.y="CIF of diabetic retinopathy", label.x="Years from registration")
```

`label.strata` is an argument for customizing labels, but when inputting a survfit object, it becomes invalid because it does not contain stratum information.
Therefore, the following code inputs the formula and data. `label.strata` is used by combining `level.strata` and `order.strata`. `level.strata` specifies the levels of the stratification variable corresponding to each label in `label.strata`. The levels specified in `level.strata` are then displayed in the graph in the order defined by `order.strata`. A graph enclosed in a square was generated, which is due to `style="FRAMED"` specification.

```{r example1-4, warning=FALSE, message=FALSE}
cifplot(Event(t,epsilon)~fruitq1, data=diabetes.complications, outcome.type="COMPETING-RISK", 
        addConfidenceInterval=FALSE, addEstimateTable=TRUE, addCensorMark=FALSE, 
        addCompetingRiskMark=TRUE, competing.risk.time=output2, 
        label.y="CIF of diabetic retinopathy", label.x="Years from registration", 
        label.strata=c("High intake","Low intake"), level.strata=c("Q2 to Q4","Q1"), 
        order.strata=c("Q1", "Q2 to Q4"), style="FRAMED")
```

By specifying `addEstimateTable=TRUE`, the risks of diabetic retinopathy (estimates for CIFs) along with their confidence interval are shown in the table at the bottom of each graph. The risk ratios at each time point may be estimated using `polyreg()` with `outcome.type="COMPETING-RISK"`. Here, no covariates are included in the nuisance model (`~1` specifies intercept only). The effect of low fruit intake `fruitq1` is estimated as an unadjusted risk ratio (`effect.measure1="RR"`) for diabetic retinopathy (event 1) and macrovascular complications (event 2) at 8 years (`time.point=8`).

```{r example2-1, warning=FALSE, message=FALSE}
output3 <- polyreg(nuisance.model=Event(t,epsilon)~1, exposure="fruitq1", 
          data=diabetes.complications, effect.measure1="RR", effect.measure2="RR", 
          time.point=8, outcome.type="COMPETING-RISK", 
          report.nuisance.parameter=TRUE)
print(output3$coefficient)
print(output3$cov)
```

The summaries of analysis results in the list of outputs (e.g. `output3$summary` below) are compatible with the `modelsummary` ecosystem, which makes it easy to create publication-ready tables in a variety of formats. A typical workflow is to pass the `summary` element of a `polyreg()` fit (or a list of such fits) to `modelsummary::msummary()`. In this case, all regression coefficients are included in summary by setting `report.nuisance.parameter = TRUE`. Model summaries can also be exponentiated to display risk ratios, odds ratios, or subdistribution hazard ratios using `exponentiate` option. The summaries can be displayed in the Viewer with customized statistics such as p-values or confidence intervals.

## Example 2. Survival analysis

The second example is time to first event analysis (`outcome.type="SURVIVAL"`) to estimate the effect on the risk of diabetic retinopathy or macrovascular complications at 8 years. In the code below, `cifplot()` is directly used to generate a survfit object internally and plot it. 

```{r example3, warning=FALSE, message=FALSE}
data(diabetes.complications)
diabetes.complications$d <- as.integer(diabetes.complications$epsilon>0)
cifplot(Event(t,d) ~ fruitq1, data=diabetes.complications, 
outcome.type="SURVIVAL", addConfidenceInterval=TRUE, addCensorMark=FALSE, 
addCompetingRiskMark=FALSE, label.y="Survival probability", 
label.x="Years from registration", label.strata=c("High intake","Low intake"))
```

The code below specifies the Richardson model on the risk of diabetic retinopathy or macrovascular complications at 8 years (outcome.type="SURVIVAL"). Dependent censoring is adjusted by stratified IPCW method (`strata='strata'`). Estimates other than the effects of exposure (e.g. intercept) are suppressed when `report.nuisance.parameter` is not specified.

```{r example4, warning=FALSE, message=FALSE}
output4 <- polyreg(nuisance.model=Event(t,d)~1, 
          exposure="fruitq1", strata="strata", data=diabetes.complications,
          effect.measure1="RR", time.point=8, outcome.type="SURVIVAL")
```

## Example 3. Adjusted competing risks analysis

The code below specifies polytomous log-odds models of both of competing events (`outcome.type="COMPETING-RISK"`). Here 15 covariates and censoring strata are specified in `nuisance.model=` and `strata=`, respectively.

```{r example5, warning=FALSE, message=FALSE}
output5 <- polyreg(nuisance.model=Event(t,epsilon)~age+sex+bmi+hba1c
          +diabetes_duration+drug_oha+drug_insulin+sbp+ldl+hdl+tg
          +current_smoker+alcohol_drinker+ltpa, 
          exposure="fruitq1", strata="strata", data=diabetes.complications,
          effect.measure1="RR", time.point=8, outcome.type="COMPETING-RISK")
```

## Example 4. Description of cumulative incidence of competing events

`cifpanel()` arranges multiple survival and CIF plots into a single, polished layout with a shared legend. It’s designed for side-by-side comparisons—e.g., event 1 vs event 2, different groupings, or different y-scales—while keeping axis ranges and styles consistent. Internally each panel is produced using the same engine as `cifcurve()`, and you can supply scalar arguments (applied to all panels) or lists to control each panel independently.

This function accepts both shared and panel-specific arguments. When a single formula is provided, the same model structure is reused for each panel, and arguments supplied as lists are applied individually to each panel. Arguments such as `code.events`, `label.strata`, or `addCensorMark` can be given as lists, where each list element corresponds to one panel. This allows flexible configuration while maintaining a concise and readable syntax.

The example below creates a 1×2 panel (`rows.columns.panel = c(1,2)`) comparing the cumulative incidence of two competing events in the same cohort, namely CIF of diabetic retinopathy in the left panel and CIF of macrovascular complications in the right panel. Both panels are stratified by `fruitq1`, and the legend is shared at the bottom. The pairs of `code.events` as a list instructs `cifpanel()` to display event 1 in the first panel and event 2 in the second panel, with event code 0 representing censoring.

```{r example7, warning=FALSE, message=FALSE}
cifpanel(
 title.panel = "A comparison of cumulative incidence of competing events",
 rows.columns.panel = c(1,2),
 formula = Event(t, epsilon) ~ fruitq1,
 data = diabetes.complications,
 outcome.type = "COMPETING-RISK",
 code.events = list(c(1,2,0), c(2,1,0)),
 label.y = c("CIF of diabetic retinopathy", "CIF of macrovascular complications"),
 label.x = "Years from registration",
 label.strata = list(c("High intake","Low intake")),
 title.plot = c("Diabetic retinopathy", "Macrovascular complications"),
 legend.position = "bottom",
 legend.collect=TRUE
)
```

Arguments specified as scalars (for example, `label.x = "Years from registration"`) are applied uniformly to all panels. Character vectors of the same length as the number of panels (for example, `label.y = c("Diabetic retinopathy", "Macrovascular complications")`) assign a different label to each panel in order. Lists provide the most flexibility, allowing each panel to have distinct settings that mirror the arguments of `cifcurve()`.

The `legend.collect = TRUE` option merges legends from all panels into a single shared legend, positioned according to `legend.position`. The arguments `title.panel`, `subtitle.panel`, `caption.panel`, and `title.plot` control the overall panel title and individual subplot titles, ensuring that multi-panel layouts remain consistent and publication-ready. 

## Additional information

### Comparison of model formula and outcome.type across cifcurve() and polyreg()

**Common ground**

-   Both functions expect a `Surv()` or `Event()` response on the left-hand side of the model formula so that follow-up time and event indicators are supplied in the same way. Internally the helpers validate non-negative times, check that the event codes match the user-specified `code.event*` arguments, and derive indicators used downstream.
-   The `outcome.type` argument is normalized by the same utility, meaning that the same set of synonyms (e.g. `"C"`, `"competing risk"`, or `"S"`) are accepted in both workflows and mapped onto the canonical labels used in the estimators.

**Key differences**

-   In `cifcurve()` the right-hand side of the formula defines the grouping variable(s) whose curves are estimated; weights (if any) are supplied separately. There is no dedicated exposure argument because all strata are handled through the formula term that is turned into a factor for plotting and estimation.
-   In `polyreg()` the `nuisance.model` formula describes only the outcome and nuisance covariates (i.e. confounders), while the exposure is passed separately through the `exposure` argument. The exposure design matrix, reference level handling, and scaling of nuisance covariates are handled explicitly for regression fitting.
-   `cifcurve()` focuses on two estimands, selecting between Kaplan–Meier and Aalen–Johansen calculations via `outcome.type = "SURVIVAL"` or `"COMPETING-RISK"`. In contrast, `polyreg()` supports additional time-constant and binomial estimands (`"PROPORTIONAL-SURVIVAL"`, `"PROPORTIONAL-COMPETING-RISK"`, `"BINOMIAL"`) and enforces time-point requirements accordingly—for example `time.point` is mandatory for survival/competing-risk contrasts, while time-constant-effect (e.g. proportional subdistribution hazards) models can either use all observed event times or a user-specified grid supplied via a (vector) `time.point`.

### Event coding conventions in package cifmodeling and CDISC ADaM ADTTE

When working with survival data, one common source of confusion arises from inconsistent event coding conventions between the standard input of `Surv()`, provided by survival package, and the CDISC ADaM ADTTE (Analysis Data Model for Time-to-Event Endpoints) used in regulatory clinical trials. With `Surv()`, event is often coded as 1 and censoring is coded as 0. By contrast, the CDISC ADaM ADTTE (commonly used in pharmaceutical submissions and SDTM-based analyses) follows the opposite convention. `CNSR = 0` means that event occurred and `CNSR = 1` represents censoring, which aligns with the broader ADaM convention (where `1 = TRUE`). 

To accommodate differences in coding conventions, one can explicitly specify event and censoring codes using the `code.event1` and `code.censoring` arguments in `cifcurve()`, `cifplot()`, `cifpanel()` and `polyreg()`. This allows users to perform accurate survival analysis using functions from the cifmodeling package while maintaining the CDISC ADaM ADTTE coding scheme. 

```{r syntax6, eval=FALSE, warning=FALSE, message=FALSE}
polyreg(Event(AVAL, CNSR) ~ ARM, data = adtte, 
        outcome.type = "SURVIVAL", code.event1=0, code.censoring=1)
```

### Processing pipeline of cifcurve()

A user-specified model formula in `Event()` or `Surv()` is first parsed by `util_read_surv()` to extract the time-to-event, the event indicator, optional weights, and any strata variable. Those components are then passed to the back-end estimators implemented in `src/calculateAJ_Rcpp.cpp`. The C++ routine supports both weighted and stratified data, so the heavy lifting for Kaplan–Meier or Aalen–Johansen estimates (and associated at-risk, event, and censor counts) happens efficiently regardless of how the input was specified. 

- The `error` argument determines how standard errors are computed. For survival data (`outcome.type = "SURVIVAL"`) it maps to Greenwood (default) or Tsiatis estimators, while for competing-risks data (`outcome.type = "COMPETING-RISK"`) `"aalen"` reproduces the classic Aalen standard errors and `"delta"` implements the delta method.

### Variance and confidence interval options for cifcurve()

**Standard errors** For standard errors of Kaplan–Meier estimator, the default is Greenwood estimator, with alternatives `"tsiatis"` for the Tsiatis estimator The Greenwood formula is a sum of terms d/(n(n-m)), where d is the number of events at a given time point, n is the number at risk, and m is the cumulative sum of events at that time. This formula is derived from a binomial argument, so extension to the weighted case is ad hoc. Tsiatis (1981) proposes a sum of terms d/(n\*n), based on a counting process argument which includes the weighted case. For confidence intervals of cumulative incidence curves, the default `"delta"` applies the delta-method, while `"aalen"` uses the Aalen variance decomposition. By default `cifcurve()` rescales the Greenwood/Tsiatis quantities so that `std.err` is reported on the probability scale; set `report.survfit.std.err = TRUE` to return the conventional log-survival standard errors from `survfit()`.

**Confidence intervals** `src/calculateAJ_Rcpp.cpp` constructs intervals on the probability scale using the requested transformation: `"arcsine-square root"`/`"arcsine"`/`"a"` (default), `"plain`, `"log"`, `"log-log"`, or `"logit"`. Passing `"none"`/`"n"` skips interval computation entirely. The function exponentiates back to the probability scale, clips bounds to [0, 1], and replaces undefined values with `NA` so that interval endpoints remain well behaved in plots and summaries.

### Processing pipeline of polyreg()

Input pre-processing mirrors the survival workflow by calling `checkInput()` to parse event codes, exposure levels, and requested estimands before `createAnalysisDataset()` builds the modeling frame and `reg_normalize_covariate()` rescales covariates by default. Starting values are obtained from the nuisance model fits via `getInitialValues()` or `getInitialValuesProportional()`, or, when necessary, from user-supplied data frame of starting values. Inverse-probability-of-censoring weights for survival and competing-risks outcomes are calculated by `calculateIPCW()`, which in turn wraps the `calculateKM()` C++ routine to obtain Kaplan-Meier estimates of censoring, while time-constant effects assemble matrices through `calculateIPCWMatrix()`.

The weighted estimating equations are then solved by `solveEstimatingEquation()`, which dispatches to specific score functions such as `estimating_equation_ipcw()`, `estimating_equation_survival()`, and `estimating_equation_proportional()` and iterates with `nleqslv`/Levenberg–Marquardt updates until the convergence criteria in `assessConvergence()` are met. Sandwich variances come from `calculateCov()` or `calculateCovSurvival()` and are rescaled back to the original covariate scale by `reg_normalize_estimate()`, while optional bootstrap intervals reuse the same solver inside the `boot` resampling loop. The resulting influence functions, weights, and fitted cumulative incidence probabilities are retained in `diagnostic.statistics` for downstream diagnostics.

### Reproducibility and conventions for polyreg()

 - If convergence warnings appear, relax/tighten tolerances or cap the parameter bound (`optim.parameter1` to `3`) and inspect `report.optim.convergence = TRUE`.
 - If necessary, try modifying other `optim.parameter`, providing user-specified initial values, or reducing the number of nuisance parameters (e.g. provide a `time.point` grid that contributes to estimation when using `"PROPORTIONAL-SURVIVAL"` and `"PROPORTIONAL-COMPETING-RISK"`).
 - Set `boot.seed` for reproducible bootstrap results.
 - Match CDISC ADaM conventions via `code.event1 = 0`, `code.censoring = 1` (and, if applicable, `code.event2` for competing events).

### Key arguments of msummary() that are helpful when reporting polyreg() results

-   `output` – controls the destination of the table (e.g., `"markdown"`, `"latex"`, `"html"`, or a file path such as `'results/cif_table.docx'`).
-   `coef_map` or `coef_rename` – renames model terms for interpretability.
-   `statistic` – which uncertainty summaries to display. Examples: `statistic = "p.value"` for p-values, `statistic = "conf.int"` for Wald confidence interval, `statistic = "t"` for t statistics, `and statistic = "z"` for z statistics, instead of standard errors (default).

You can also supply multiple model summaries as a named list to compare estimates across specifications in a single table. See `vignette('modelsummary')` for a comprehensive overview of additional layout and styling options.

## Full lists of arguments

### cifcurve()

-   `formula` A model formula specifying the time-to-event outcome on the left-hand side (typically `Event(time, status)` or `survfit::Surv(time, status)`) and, optionally, a stratification variable on the right-hand side. Unlike `cifplot()`, this function does not accept a fitted `survfit` object. 
-   `data` A data frame containing variables in `formula`.
-   `weights` Optional name of the weight variable in `data`. Weights must be nonnegative.
-   `subset.condition` Optional character expression to subset `data` before analysis.
-   `na.action` A function specifying the action to take on missing values. The default is `na.omit`.
-   `outcome.type` Character string specifying the type of time-to-event outcome. One of `"SURVIVAL"` (Kaplan–Meier type) or `"COMPETING-RISK"` (Aalen–Johansen type). If `NULL` (default), the function automatically infers the outcome type from the data: if the event variable has more than two unique levels, `"COMPETING-RISK"` is assumed; otherwise, `"SURVIVAL"` is used. You can also use abbreviations such as `"S"` or `"C"`. Mixed or ambiguous inputs (e.g., `c("S", "C")`) trigger automatic detection based on the event coding in `data`.
-   `code.event1` Integer code of the event of interest (default `1`).
-   `code.event2` Integer code of the competing event (default `2`).
-   `code.censoring` Integer code of censoring (default `0`).
-   `error` Character specifying the method for variance and standard error used internally. For `"SURVIVAL"` typically `"greenwood"`. For `"COMPETING-RISK"` pass options supported by `calculateAalenDeltaSE()` (`"aalen"`, `"delta"`, `"none"`).
-   `conf.type` Character transformation for CI on the probability scale (default `"arcsine-square root"`).
-   `conf.int` numeric two-sided confidence level (default `0.95`).
-   `return_if` Logical. When `TRUE` and `engine = "calculateAJ_Rcpp"`, the influence function is also computed and returned (default `FALSE`).
-   `report.survfit.std.err` Logical. If `TRUE`, report SE on the log-survival scale (survfit's convention). Otherwise SE is on the probability scale.
-   `engine` Character. One of `"auto"`, `"calculateKM"`, or `"calculateAJ_Rcpp"` (default `"calculateAJ_Rcpp"`). 
-   `prob.bound` Numeric lower bound used to internally truncate probabilities away from 0 and 1 (default `1e-7`).



### cifplot()

-   `formula_or_fit` A model formula or a survfit object. Note: When a formula is supplied, the left-hand side must be `Event(time, status)` or `survfit::Surv(time, status)`.
-   `data` A data frame containing variables in `formula`.
-   `weights` Optional name of the weight variable in `data`. Weights must be nonnegative.
-   `subset.condition` Optional character expression to subset `data` before analysis.
-   `na.action` A function specifying the action to take on missing values. The default is `na.omit`.
-   `outcome.type` Character string specifying the type of time-to-event outcome. One of `"SURVIVAL"` (Kaplan–Meier type) or `"COMPETING-RISK"` (Aalen–Johansen type). If `NULL` (default), the function automatically infers the outcome type from the data: if the event variable has more than two unique levels, `"COMPETING-RISK"` is assumed; otherwise, `"SURVIVAL"` is used. You can also use abbreviations such as `"S"` or `"C"`. Mixed or ambiguous inputs (e.g., `c("S", "C")`) trigger automatic detection based on the event coding in `data`.
-   `code.event1` Integer code of the event of interest (default `1`).
-   `code.event2` Integer code of the competing event (default `2`).
-   `code.censoring` Integer code of censoring (default `0`).
-   `error` Character specifying the method for variance and standard error used internally. For `"SURVIVAL"` typically `"greenwood"`. For `"COMPETING-RISK"` pass options supported by `calculateAalenDeltaSE()` (`"aalen"`, `"delta"`, `"none"`).
-   `conf.type` Character transformation for CI on the probability scale (default `"arcsine-square root"`).
-   `conf.int` numeric two-sided confidence level (default `0.95`).
-   `type.y` Optional vector/list per panel: `"surv"` or `"risk"` (display `1 - survival` i.e. CIF).
-   `label.x` Character x-axis labels (default `"Time"`).
-   `label.y` Character y-axis labels (default internally set to `"Survival"` or `"Cumulative incidence"`).
-   `label.strata` Character vector of labels for strata.
-   `order.strata` Optional ordering of strata levels. When `printEachVar = TRUE`, supply a named list `list(var = c("L1","L2",...))` for each RHS variable; unmatched levels are dropped. When `printEachVar = FALSE`, supply a character vector `c("L1","L2",...)` that specifies the display order (legend/risktable) of the single stratification factor. Levels not listed are dropped. If `label.strata` is a named vector, its names must match the (re-ordered) levels.
-   `limits.x` Numeric length-2 vectors for axis limits. If `NULL` it is internally set to `c(0,max(out_readSurv$t))`.
-   `limits.y` Numeric length-2 vectors for axis limits. If `NULL` it is internally set to `c(0,1)`.
-   `breaks.x` Numeric vectors for axis breaks (default `NULL`).
-   `breaks.y` Numeric vectors for axis breaks (default `NULL`).
-   `use_coord_cartesian` Logical specify use of `coord_cartesian()` (default `FALSE`).
-   `addConfidenceInterval` Logical add `add_confidence_interval()` to plot. It calls `geom_ribbon()` (default `TRUE`).
-   `addRiskTable` Logical add `add_risktable(risktable_stats="n.risk")` to plot (default `TRUE`).
-   `addEstimateTable` Logical add `add_risktable(risktable_stats="estimate (conf.low, conf.high)")` to plot (default `FALSE`).
-   `symbol.risktable` Character specifying the symbol used in the risk table to denote strata: `"square"`, `"circle"`, or `"triangle"` (default `"square"`).
-   `font.size.risktable` Numeric font size for texts in risk / estimate tables (default `3`).
-   `addCensorMark` Logical add `add_censor_mark()` to plot. It calls `geom_point()` (default `TRUE`).
-   `shape.censor.mark` Integer point shape for censor marks (default `3`).
-   `size.censor.mark` Numeric point size for censor marks (default `2`).
-   `addCompetingRiskMark` Logical add time marks to describe event2 specified by `Event()`, usually the competing events. It calls `geom_point()` (default `TRUE`).
-   `competing.risk.time` Named list of numeric vectors (names must be mapped to strata labels).
-   `shape.competing.risk.mark` Integer point shape for competing-risk marks (default `16`).
-   `size.competing.risk.mark` Numeric point size for competing-risk marks (default `2`).
-   `addIntercurrentEventMark` Logical overlay user-specified time marks per strata calls `geom_point()` (default `TRUE`).
-   `intercurrent.event.time` Named list of numeric vectors (names must be mapped to strata labels).
-   `shape.intercurrent.event.mark` Integer point shape for intercurrent-event marks (default `1`).
-   `size.intercurrent.event.mark` Numeric point size for intercurrent-event marks (default `2`).
-   `addQuantileLine` Logical add `add_quantile()` to plot. It calls `geom_segment()` (default `TRUE`).
-   `quantile` Numeric specify quantile for `add_quantile()` (default `0.5`).
-   `printEachEvent` Logical. Explicit panel mode. If `TRUE` and `outcome.type == "COMPETING-RISK"`, `cifplot()` internally calls `cifpanel()` to display two event-specific CIFs side-by-side (event 1 and event 2) using reversed `code.events`. Ignored for non-competing-risk outcomes.
-   `printCensoring` Logical. Explicit panel mode. If `TRUE` and `outcome.type == "SURVIVAL"`, `cifplot()` internally calls `cifpanel()` to display KM-type curves for (event, censor) and (censor, event) so that censoring patterns can be inspected.
-   `printEachVar` Logical. Explicit panel mode. If `TRUE` and the right-hand side of the formula has multiple covariates (e.g. `~ a + b + c`), the function produces a panel where each variable in RHS is used once as the stratification factor.
-   `panel.mode` Character specifying Automatic panel mode. If `"auto"` and none of `printEachVar`, `printEachEvent`, `printCensoring` has been set to `TRUE`, the function chooses a suitable panel mode automatically: (i) if the formula RHS has 2+ variables, it behaves like `printEachVar = TRUE`; (ii) otherwise, if `outcome.type == "COMPETING-RISK"`, it behaves like `printEachEvent = TRUE`; (iii) otherwise, if `outcome.type == "SURVIVAL"`, it behaves like `printCensoring = TRUE`. If a panel mode is explicitly specified, `panel.mode` is ignored.
-   `rows.columns.panel` Optional integer vector `c(nrow, ncol)` controlling the panel layout. If `NULL`, an automatic layout is used.
-   `style` Character plot theme controls (default `"CLASSIC"`).
-   `palette` Optional character vector specify color palette, e.g. `palette=c("blue", "cyan", "navy", "green")` (default `NULL`).
-   `font.family` Character plot theme controls (e.g. `"sans"`, `"serif"`, and `"mono"`. default `"sans"`).
-   `font.size` Integer plot theme controls (default `12`).
-   `legend.position` Character specify position of legend: `"top"`, `"right"`, `"bottom"`, `"left"`, or `"none"` (default `"top"`).
-   `filename.ggsave` Character save the `ggsurvfit` plot with the path and name specified.
-   `width.ggsave` Numeric specify width of the `ggsurvfit` plot.
-   `height.ggsave` Numeric specify height of the `ggsurvfit` plot.
-   `dpi.ggsave` Numeric specify dpi of the `ggsurvfit` plot.

### cifpanel()

-   `formula` A model formula specifying the time-to-event outcome on the left-hand side (typically `Event(time, status)` or `survfit::Surv(time, status)`) and, optionally, a stratification variable on the right-hand side. Unlike `cifplot()`, this function does not accept a fitted `survfit` object. 
-   `formulas` A list of formulas (one per panel). If provided, overrides `formula`.
-   `data` A data frame containing variables in `formula`.
-   `weights` Optional name of the weight variable in `data`. Weights must be nonnegative.
-   `subset.condition` Optional character expression to subset `data` before analysis.
-   `na.action` A function specifying the action to take on missing values. The default is `na.omit`.
-   `outcome.type` Character string specifying the type of time-to-event outcome. One of `"SURVIVAL"` (Kaplan–Meier type) or `"COMPETING-RISK"` (Aalen–Johansen type). If `NULL` (default), the function automatically infers the outcome type from the data: if the event variable has more than two unique levels, `"COMPETING-RISK"` is assumed; otherwise, `"SURVIVAL"` is used. You can also use abbreviations such as `"S"` or `"C"`. Mixed or ambiguous inputs (e.g., `c("S", "C")`) trigger automatic detection based on the event coding in `data`.
-   `code.event1` Integer code of the event of interest (default `1`).
-   `code.event2` Integer code of the competing event (default `2`).
-   `code.censoring` Integer code of censoring (default `0`).
-   `code.events` Optional numeric length-3 vector `c(event1, event2, censoring)`. When supplied, it overrides `code.event1`, `code.event2`, and `code.censoring` (primarily used when `cifpanel()` is called or when `printEachEvent = TRUE`).
-   `error` Character specifying the method for variance and standard error used internally. For `"SURVIVAL"` typically `"greenwood"`. For `"COMPETING-RISK"` pass options supported by `calculateAalenDeltaSE()` (`"aalen"`, `"delta"`, `"none"`).
-   `conf.type` Character transformation for CI on the probability scale (default `"arcsine-square root"`).
-   `conf.int` numeric two-sided confidence level (default `0.95`).
-   `type.y` Optional vector/list per panel: `"surv"` or `"risk"` (display `1 - survival` i.e. CIF).
-   `label.x`, `label.y` Optional vectors/lists of axis labels per panel.
-   `label.strata` Optional list of character vectors for legend labels per panel (passed to `cifplot()`).
-   `order.strata` Optional list of character vectors for ordering labels per panel (passed to `cifplot()`).
-   `limits.x`, `limits.y` Optional vectors/lists of numeric length-2 axis limits per panel.
-   `breaks.x`, `breaks.y` Optional vectors/lists of axis breaks per panel (forwarded to `breaks.x` / `breaks.y` in `cifplot()`).
-   `addConfidenceInterval`, `addCensorMark`, `addCompetingRiskMark`, `addIntercurrentEventMark`, `addQuantileLine` Optional logical vectors/lists per panel to toggle features in `cifplot()`. If `NULL`, sensible defaults are used (CI/Censor on; others off).
-   `rows.columns.panel` Integer vector `c(nrow, ncol)` specifying the grid size.
-   `title.panel`, `subtitle.panel`, `caption.panel` Optional strings for panel annotation.
-   `tag.panel` Passed to `patchwork::plot_annotation(tag_levels = ...)`.
-   `title.plot` Optional length-2 character vector, titles for base/inset plots when `inset.panel = TRUE`.
-   `legend.position` Position of legends: `"top"`, `"right"`, `"bottom"`, `"left"`, or `"none"`.
-   `legend.collect` If `TRUE` (grid mode), collect legends across subplots.
-   `inset.panel` If `TRUE`, place the second plot as an inset over the first.
-   `inset.left`, `inset.bottom`, `inset.right`, `inset.top` Numeric positions (0–1) of the inset box.
-   `inset.align.to` One of `"panel"`, `"plot"`, or `"full"`.
-   `inset.legend.position` Legend position for the inset plot (e.g., `"none"`).
-   `filename.ggsave` Character save the composed panel with the path and name specified.
-   `width.ggsave` Numeric specify width of the composed panel.
-   `height.ggsave` Numeric specify height of the composed panel.
-   `dpi.ggsave` Numeric specify dpi of the composed panel.
-   `...` Additional arguments forwarded to `cifplot` (e.g., `style`, `font.family`, `font.size`, etc.). Panel-wise overrides provided via explicit arguments take precedence over `...`.


### polyreg()

-   `nuisance.model` A `formula` describing the outcome and nuisance covariates, excluding the exposure of interest. The left-hand side must be `Event(time, status)` or `survfit::Surv(time, status)`.
-   `exposure` A character string giving the name of the categorical exposure variable in `data`.
-   `strata` Optional character string with the name of the stratification variable used to adjust for dependent censoring (default `NULL`).
-   `data` A data frame containing the outcome, exposure and nuisance covariates referenced by `nuisance.model`.
-   `subset.condition` Optional expression (as a character string) defining a subset of `data` to analyze (default `NULL`).
-   `code.event1` Integer code of the event of interest (default `1`).
-   `code.event2` Integer code of the competing event (default `2`).
-   `code.censoring` Integer code of censoring (default `0`).
-   `code.exposure.ref` Integer code identifying the reference exposure category (default `0`).
-   `effect.measure1` Character string specifying the effect measure for the primary event. Supported values are `"RR"`, `"OR"` and `"SHR"`.
-   `effect.measure2` Character string specifying the effect measure for the competing event. Supported values are `"RR"`, `"OR"` and `"SHR"`.
-   `time.point` Numeric time point at which the exposure effect is evaluated. Required for survival and competing risk analyses.
-   `outcome.type` Character string selecting the outcome type. Valid values are `"COMPETING-RISK"`, `"SURVIVAL"`, `"BINOMIAL"`, `"PROPORTIONAL-SURVIVAL"` and `"PROPORTIONAL-COMPETING-RISK"`. Defaults to `"COMPETING-RISK"`. If `NULL` (default), the function automatically infers the outcome type from the data: if the event variable has more than two unique levels,
`"COMPETING-RISK"` is assumed; otherwise, `"SURVIVAL"` is used. You can also use abbreviations such as `"S"` or `"C"`. Mixed or ambiguous inputs (e.g., `c("S", "C")`) trigger automatic detection based on the event coding in `data`.
-   `conf.int` numeric two-sided confidence level (default `0.95`).
-   `report.nuisance.parameter` Logical if `TRUE`, the returned object includes estimates of the nuisance model parameters (default `FALSE`).
-   `report.optim.convergence` Logical if `TRUE`, optimization convergence summaries are returned (default `FALSE`).
-   `report.sandwich.conf` Logical or `NULL`. When `TRUE`, confidence intervals based on sandwich variance are computed. When `FALSE`, they are omitted (default `TRUE`).
-   `report.boot.conf` Logical or `NULL`. When `TRUE`, bootstrap confidence intervals are computed. When `FALSE`, they are omitted. If `NULL`, the function chooses based on `outcome.type` (default `NULL`).
-   `boot.bca` Logical indicating the bootstrap confidence interval method. Use `TRUE` for bias-corrected and accelerated intervals or `FALSE` for the normal approximation (default to `FALSE`).
-   `boot.replications` Integer giving the number of bootstrap replications (default to `200`).
-   `boot.seed` Numeric seed used for resampling of bootstrap.
-   `nleqslv.method` Character string specifying the solver used in `nleqslv()`. Available choices are `"Broyden"` and `"Newton"`.
-   `optim.parameter1` Numeric tolerance for convergence of the outer loop (default `1e-6`).
-   `optim.parameter2` Numeric tolerance for convergence of the inner loop (default `1e-6`).
-   `optim.parameter3` Numeric constraint on the absolute value of parameters (default `100`).
-   `optim.parameter4` Integer maximum number of outer loop iterations (default `50`).
-   `optim.parameter5` Integer maximum number of `nleqslv()` iterations per outer iteration (default `50`).
-   `optim.parameter6` Integer maximum number of iterations for the Levenberg-Marquardt routine (default `50`).
-   `optim.parameter7` Numeric convergence tolerance for the Levenberg-Marquardt routine (default `1e-10`).
-   `optim.parameter8` Numeric tolerance for updating the Hessian in the Levenberg-Marquardt routine (default `1e-6`).
-   `optim.parameter9` Numeric starting value for the Levenberg-Marquardt damping parameter lambda (default `1e-6`).
-   `optim.parameter10` Numeric upper bound for lambda in the Levenberg-Marquardt routine (default `40`).
-   `optim.parameter11` Numeric lower bound for lambda in the Levenberg-Marquardt routine (default `0.025`).
-   `optim.parameter12` Numeric multiplicative increment applied to lambda when the Levenberg-Marquardt step is successful (default to `2`).
-   `optim.parameter13` Numeric multiplicative decrement applied to lambda when the Levenberg-Marquardt step is unsuccessful (default `0.5`).
-   `data.initial.values` Optional data frame providing starting values for the optimization (default `NULL`).
-   `normalize.covariate` Logical indicating whether covariates should be centered and scaled prior to optimization (default `TRUE`).
-   `terminate.time.point` Logical indicating whether time points that contribute estimation are terminated by min of max follow-up times of each exposure level (default `TRUE`).
-   `prob.bound` Numeric lower bound used to internally truncate probabilities away from 0 and 1 (default `1e-7`).
