---
title: "Main functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Main functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup02, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.path   = "fig/function-",
  fig.ext    = "png",
  fig.width  = 6,
  fig.height = 4,
  dpi        = 300,
  fig.retina = 2,
  fig.align  = "center",
  out.width  = "70%"
)

if (requireNamespace("cifmodeling", quietly = TRUE)) {
  library(cifmodeling)
} else {
  stop("Package 'cifmodeling' is not available during vignette build.")
}
library(nleqslv)
library(boot)
library(ggplot2)
library(ggsurvfit)
library(patchwork)

data("diabetes.complications", package = "cifmodeling")
set.seed(1)
```

<!-- badges: start --> <!-- badges: end -->

### cifplot()

This function produces the **Kaplan–Meier survival** or **Aalen–Johansen cumulative incidence** curve from a unified formula + data interface. It auto-labels axes based on `outcome.type` and `type.y` (`"risk"` for 1-survival or CIF), can add censoring/competing-risk/intercurrent-event marks, and returns a regular ggplot object (compatible with + and %+%). You may also pass a survfit-compatible object directly.

**Typical use cases**

-   Draw one survival/CIF curve set by exposure groups (e.g., treatment vs control).
-   Call `cifpanel()` with a simplified code to create a panel displaying plots of multiple stratified survival/CIF curves or CIF curves for each event type.
-   Add CIs and censor/competing-risk/intercurrent-event marks.
-   Add a risk table to display the number at risk or the estimated survival probabilities or CIFs and CIs at each point in time.

**Key arguments shared with cifcurve()**

-   **Outcome type and estimator**
    -   `formula` A model formula specifying the outcome (`Event(time, status)` or `Surv(time, status)`) on the left-hand side (LHS) and, optionally, a stratification variable on the right-hand side (RHS). 
    -   `outcome.type = "survival"` → Kaplan–Meier estimator
    -   `outcome.type = "competing-risk"` → Aalen–Johansen estimator
-   **Confidence intervals**
    -   `conf.int` sets the two-sided level (default 0.95)\
    -   `conf.type` chooses the transformation (`"arcsine-square root"`, `"plain"`, `"log"`, `"log-log"`, `"logit"`, or `"none"`)
    -   `error` chooses the estimator for SE (`"greenwood"`, `"tsiatis"` or `"if"` for survival curves and `"delta"` or `"aalen"` or `"if"` for CIFs)

**Key arguments for cifplot()**

-   **Data visualization**
    -   `add.conf` adds CIs on the ggplot2-based plot
    -   `add.competing.risk.mark` and `add.intercurrent.event.mark` add symbols to describe competing risks or intercurrent events in addition to conventional censoring marks with `add.censor.mark`
    -   `add.risktable` adds numbers at risk
    -   `add.estimate.table` adds time-by-time estimates and CIs
    -   `add.quantile` add a reference line at a chosen quantile level

-   **Plot customization**
    -   `type.y` chooses y-axis (`"surv"` for survival and `"risk"` for 1-survival/CIF)
    -   `limits.x`, `limits.y`, `breaks.x`, `breaks.y` numeric vectors for axis control
    -   `style` specifies the appearance of plot (`"classic"`, `"bold"`, `"framed"`, `"grid"`, `"gray"` or `"ggsurvfit"`)
    -   `palette` specifies color of each curve (e.g. `palette=c("blue1", "cyan3", "navy", "deepskyblue3"))`)

-   **Panel display**
    -   `panel.per.variable` produces multiple survival/CIF curves per stratification variable specified in the formula
    -   `panel.per.event` produces CIF curves for each event type
    -   `panel.censoring` produces the Kaplan–Meier curves for (event, censor) and (censor, event) so that censoring patterns can be inspected
    -   `panel.mode` uses automatic panel mode

**Return**

-   A **cifplot** object (list) with at least: 
    -   `plot` the **ggplot** object
    -   Additional metadata used by `cifpanel()` and helper functions

**Under the hood: cifcurve()**

`cifplot()` is a streamlined, opinionated wrapper around `cifcurve()`, which calculates the Kaplan–Meier estimator and the Aalen–Johansen estimator. `cifcurve()` returns a survfit-compatible object, enabling an independent use of standard methods such as:

-   `summary()` — time-by-time estimates with SEs and CIs
-   `plot()` — base R stepwise survival/CIF curves
-   `mean()` — restricted mean survival estimates with CIs
-   `quantile()` — quantile estimates with CIs

Furthermore, the cifmodeling package provides a workflow for a **ggplot2-based visualization** based on the survfit-compatible objects by passing them to `cifplot()` and `cifpanel`. 

### cifpanel()

`cifpanel()` arranges multiple `cifplot()`/ggplot (or `cifcurve()` results rendered via `cifplot()`) into a unified layout with consistent scales, shared legend, and synchronized themes. The inset feature allows you to display one plot inside another.

**Typical use cases**
-   Compare CIF (event 1) vs CIF (event 2) in a 1×2 layout.

-   Compare survival/CIF curves across strata with a shared legend and matched axes.

-   Display a plot with an enlarged y-axis inside a full-scale plot.

**Key arguments**

-   `formula` or `formulas` — one formula or a list of formulas; each entry creates a panel.
-   `data`, `outcome.type`, `code.events`, `type.y` — recycled across panels unless a list is supplied for per-panel control.
-   `rows.columns.panel` — selects grid layout by c(rows, cols).
-   `inset.panel` — selects inset layout.
-   `title.panel`, `subtitle.panel`, `caption.panel`, `title.plot` — overall titles and captions.
-   `tag.panel` — panel tag style (e.g., "A", "a", "1").
-   `label.x`, `label.y`, `limits.x`, `limits.y`, `breaks.x`, `breaks.y` — shared axis control unless a list is supplied for per-panel control. 

**Return**

-   A **cifpanel** object (list) with at least: 
    -   `patchwork` the **patchwork** object (still ggplot-compatible)
    -   `plots` a list of **ggplot** objects in the `patchwork`
    -   Additional metadata

**An example of usage**

In the first example, `cifplot()` called `cifpanel()` to combine multiple plots. `cifpanel()` is a function for creating a panel with multiple plots like this. `cifpanel()` can also display one plot inside another. The cumulative incidence probability for macrovascular complications is low, making it difficult to discern differences between groups. The following code uses `inset.panel=` to display plot `output2`, with an enlarged vertical axis, inside plot `output1`. The position and size of the zoomed-in-view panel is specified by `inset.left`, `inset.bottom`, `inset.right`, and `inset.top`. 

```{r example02-1, fig.cap="Zoomed-in-view panel using cifplot() and cifpanel()", warning=FALSE, message=FALSE}
output1 <- cifplot(Event(t,epsilon) ~ fruitq,
                   data=diabetes.complications,
                   outcome.type="competing-risk",
                   code.event1=2,
                   code.event2=1,
                   add.conf=FALSE,
                   add.risktable = FALSE,
                   label.y="CIF of macrovascular complications",
                   label.x="Years from registration")
output2 <- cifplot(Event(t,epsilon) ~ fruitq,
                   data=diabetes.complications,
                   outcome.type="competing-risk",
                   code.event1=2,
                   code.event2=1,
                   add.conf = FALSE,
                   add.risktable = FALSE,
                   label.y="",
                   label.x="",
                   limits.y=c(0,0.15))
output3 <- list(a = output1$plot, b = output2$plot)
output4 <- cifpanel(plots=output3,
         title.plot=c("Fruit intake and macrovascular complications", "Zoomed-in view"),
         inset.panel=TRUE,
         inset.left=0.40, inset.bottom=0.45,
         inset.right=1.00, inset.top=0.95,
         inset.legend.position="none",
         legend.position="bottom")
print(output4)
```

The code below, in which a formula and data is given directly, produces the same output as above. 

```{r example02-2, eval=FALSE, warning=FALSE, message=FALSE}
cifpanel(
 title.plot = c("Fruit intake and macrovascular complications", "Zoomed-in view"),
 inset.panel = TRUE,
 formula = Event(t, epsilon) ~ fruitq,
 data = diabetes.complications,
 outcome.type = "competing-risk",
 code.events = list(c(2,1,0), c(2,1,0)),
 label.y = c("CIF of macrovascular complications", ""),
 label.x = c("Years from registration", ""),
 limits.y = list(c(0,1), c(0,0.15)),
 inset.left = 0.40, inset.bottom = 0.45,
 inset.right = 1.00, inset.top = 0.95,
 inset.legend.position = "none",
 legend.position = "bottom", 
 add.conf = FALSE
)
```

## polyreg()

This function implements **log odds product modeling** of CIFs at user-specified time points, focusing on multiplicative effects of a categorical exposure, or constant effects over time like Cox regression and Fine-Gray models. It estimates multiplicative effects such as **risk ratios**, **odds ratios**, or **subdistribution hazard ratios**, while ensuring that the probabilities across competing events sum to one. This is achieved through **reparameterization using polytomous log odds products**, which fits so-called effect-measure models and nuisance models on multiple competing events simultaneously. Additionally, `polyreg()` supports direct binomial regression for a survival outcome and the Richardson model for a binomial outcome, both of which use log odds products.

The function follows the familiar **formula + data** syntax with `Event()` or `Surv()` and outputs tidy results, including point estimates, SEs, CIs, and p-values. Its results can be easily summarized with `summary()` or combined with tools such as **modelsummary** or **broom** for reporting.

**Key arguments**

-   `nuisance.model` — a formula describing the outcome and nuisance covariates, excluding the exposure of interest. 
-   `exposure` — specifies the exposure variable. This argument accepts one binary or categorical variable.
-   `effect.measure1` and `effect.measure2` — specifies the effect measures for event1 and event2 (`"RR"`, `"OR"` or `"SHR"`). 
-   `outcome.type` selects the outcome type (`"competing-risk"`, `"survival"`, `"binomial"`, `"proportional-survival"` or `"proportional-competing-risk"`). 
-   `time.point` — specifies time points at which the exposure effect is evaluated. Required for `"competing-risk"` and `"survival"` outcomes.
-   `strata` — specifies the stratification variable used for the IPCW adjustment for dependent censoring. 

**Return**

-   `coefficient` — regression coefficients

-   `cov` — variance-covariance matrix for regression coefficients

-   `diagnostic.statistics` — a data frame containing inverse probability weights, influence functions, and predicted potential outcomes

-   `summary` — a summary of estimated exposure effects

### An example of usage

```{r example02-3, eval=FALSE, warning=FALSE, message=FALSE}
output <- polyreg(nuisance.model=Event(t,epsilon)~1, exposure="fruitq", 
          data=diabetes.complications, effect.measure1="RR", 
          effect.measure2="RR", time.point=8, outcome.type="competing-risk")
```
